\chapter{Continuous Functions and Uniform Convergence} 

3"1. The dependence of one complex number on another.

The problems with which Analysis is mainly occupied relate to the
dependence of one complex number on another. If z and are two complex
numbers, so connected that, if z is given any one of a certain set of
values, corresponding values of can be determined, e.g. if is the
square of z, or if = 1 when z is real and = for all other values of z,
then is said to be a function of z.

This dependence must not be confused with the most important case of
it, which will be explained later under the title o analytic
functionality.

If f is a real function of a real variable z, then the relation
between ( and 2, which may be written

can be visualised by a curve in a plane, namely the locus of a point
whose coordinates referred to rectangular axes in the plane are (s,
(). No such simple and convenient geometrical method can be found for
visualising an equation

considered as defining the dependence of one complex number f = | +
i'7 on another complex number z = x + i)/. A representation strictly
analogous to the one already given for real variables would require
four-dimensional space, since the number of variables , fj, X, y is
now four.

One suggestion (made by Lie and Weierstrass) is to use a
doubly-manifold system of lines in the quadruply-manifold totality of
lines in three-dimensional space.

Another suggestion is to represent and;; separately by means of
surfaces

A third suggestion, due to Heffter*, is to write

then draw the surface r = r x, y) — which may be called the
modular-surface of the function— and on it to express the values of 6
by surface-markings. It might be possible to modify this suggestion in
various ways by representing 6 by curves drawn on the surface r=r (.r,
y).

3'2. Continuity of functions of real variables.

The reader will have a general idea (derived from the graphical
represen- tation of functions of a real variable) as to what is meant
by continuity.

* Zeitschrift fur Math, tind Phys. xlix. (1899), p. 235.



42 THE PROCESSES OF ANALYSIS [cHAP. Ill

We now have to give a precise definition which shall embody this vague
idea.

Jjetf x) be a function of x defined when a x b.

Let x be such that a x b. If there exists a number I such that,
corresponding to an arbitrary positive number e, we can find a
positive number rj such that

\ f a:)-l\ < e,

whenever \ x — Xi\ < rj, x x , and a x b, then I is called the limit
of /(a;)

It may happen that we can find a number 1+ (even when I does not
exist) such that \ f(x) — l+\ < € when x- < x < Xi + rj. We call Z+
the limit of f x) when X approaches x- from the right and denote it
by/(a;i + 0); in a similar manner we define f x — 0) if it exists.

If f(xi + 0), f xi), f x- — 0) all exist and are equal, we say that
/(a-) is continuous at x; so that \ if x) is continuous at x- , then,
given e, we can find 7] such that

\ f x) -f(x,) I < e,

whenever \ x — Xi\ < rj and a x h.

If 1+ and l\ exist but are unequal, f x) is said to have an ordinary
discontinuity* at Xi] and if 1+ = /\ 4=/( i), f ) is said to have a
removable discontinuity at x- .

lif x) is a complex function of a real variable, and \ if x) = g(x) +
i h (x) where g (x) and h (x) are real, the continuity of f(x) at x
implies the con- tinuity of g (x) and of A (x). For when \ f(x) —f x \
< e, then | g (x) —g(xi) \ < e and I h (x) — h xi)\ < e; and the
result stated is obvious.

Example. From 2*2 examples 1 and 2 deduce that if f x) and cf) (x) are
con- tinuous at Xi , so are f x) + x), f x) x x) and, if (.rj) =t= 0,
f x)/(l) (.*•).

The popular idea of continuity, so far as it relates to a i-eal
variable f x) depending on another real variable x, is somewhat
different from that just considered, and may perhaps best be expressed
by the statement "The function f x) is said to depend con- tinuously
on X if, as x passes through the set of all values intermediate
between any two adjacent values Xi and X2, f x) passes through the set
of all values intermediate between the corresponding values /( j)
and/(.r2)."

The question thus arises, how far this popular definition is
equivalent to the precise definition given above.

Cauchy shewed that if a real function f x), of a real variable . •,
satisfies the precise definition, then it also satisfies what we have
called the popular definition; this result

* If a function is said to have ordinary discontinuities at certain
points of an interval it is implied that it is continuous at all other
points of the interval.



3*2 1] CONTINUOUS FUNCTIONS AND UNIFORM CONVERGENCE 43

vill be proved in 3"63. But the converse is not true, as was shewn by
Darboux. This fact may be iUustrated by the following example*.

Between x= — I and x= +1 (except at x=0), let f x) = sin —; and
let/(0)=0.

It can then be proved that/ .r) depends continuously on x near .r=0,
in the sense of the popular definition, but is not continuous at a; =
in the sense of the precise definition.

Example. If f x) be defined and be an increasing function in the range
(a, b), the limits /(.r + 0) exist at all points in the interior of
the range.

[If f x) be an increasing function, a section of rational numbers can
be found such that, if a, A be any members of its Z-cla.ss and its
-clas.s, a<f x + h) for evei-y positive value of h and A ' f x- h) for
some po.sitive value of h. The number defined by this section is/(a; +
0).]

3"21. Simple curves. Continua.

Let X and y be two real functions of a real variable t which are
continuous for every value of t such that a- t %h. We denote the
dependence of x and y on t by writing

.' = •(0. 3/ = i/(0- a. t%h)

The functions x (t), y (t) are supposed to be such that they do not
assume the same pair of values for any two different values of t in
the range a < t < b.

Then the set of points with coordinates (x, y) corresponding to these
values of t is called a simple curve. If

X (a) = X h), y (a) = y (b), the simple curve is said to be closed.

Example. The circle x + 7/-= 1 is a simple closed curve; for we may
write t

x=coiit, y = smt. (0 27r)

A two-dimensional continuum is a set of points in a plane possessing
the following two properties :

(i) If (x, y) be the Cartesian coordinates of any point of it, a
positive number 8 (depending on x and y) can be found such that every
point whose distance from x, y) is less than S belongs to the set.

(ii) Any two points of the set can be joined by a simple curve
consisting entirel '' of points of the set.

Example. The points for which .r-+3/-<l form a continuum. For if P be
any point inside the unit circle such that OP=r<\, we may take 8=1-?-;
and any two points inside the circle may be joined by a straight line
lying wholly inside the circle.

The following two theorems | will be assumed in this work; simple
cases of them appear ob\ dous from geometrical intuitions and,
generally, theorems of a similar nature will be taken for granted, as
formal proofs are usually extremely long and difficult.

* Due to Mansion, Mathesis, (2) xix. (1899), pp. 129-131.

t For a proof that the sine and cosine are continuous functions, see
the Appendix, § A-41. + Formal proofs will be found in Watson's
Complex Integration and Cauchy's Theorem. (Cambridge Math. Tracts, No.
15.)



44 THE PROCESSES OF ANALYSIS [CHAP. Ill

(I) A simple closed curve divides the plane into two continua (the '
interior ' and the ' exterior '). '

(II) If P be a point on the curve and Q be a point not on the curve,
the angle between QP and Ox increases by + 27r or by zero, as P
describes the curve, according as Q is an interior point or an
exterior point. If the increase is + 27r, P is said to describe the
curve ' counterclockwise.'

A continuum formed by the interior of a simple curve is sometimes
called an open two-dimensional region, or briefly an open region, and
the curve is called its boundary; such a continuum with its boundary
is then called a closed two-dimensional region, or briefly a closed
region or domain.

A simple curve is sometimes called a closed one-dimensional region; a
simple curve with its end-points omitted is then called an open
one-dimensional region.

3"22. Continuous functions of complex variables.

Lety"( ) be a function of defined at all points of a closed region
(one- or two-dimensional) in the Argand diagram, and let z be a point
of the region.

Then f z) is said to be continuous at z , if given any positive number
e, we can find a corresponding positive number 77 such that

\ f z)-f z,)\ < e, whenever \ z — Zx\ < r) and 2- is a point of the
region.

3'3. Series of variable tei-ms. Uniformity of convergence. Consider
the series

  1 + a;2 (1+ ieO' (1 + T

This series converges absolutely (§ 2*33) for all real values of x. If
8n ( ) be the sum of n terms, then ' >•

' " ">= +" -(TT ' /

and so lim Sn x) = \ \ x'; xrf 0)

but Sn 0) = 0, and therefore lim Sn (0) = 0.

M-*-00

Consequently, although the series is an absolutely convergent series
of continuous functions of x, the sum is a discontinuous function of
x. We naturally enquire the reason of this rather remarkable
phenomenon, which was investigated in 1841-1848 by Stokes*, Seidelf
and Weierstrassj, who shewed that it cannot occur except in connexion
with another phenomenon, that of non-uniform convergence, which will
now be explained.

* Cnmb. Phil. Trans, viii. (1847), pp. 533-583. [Collected Papers, i.
pp. 236-313.] t Mi'mchener Abhandlungen, v. (1848), p. 381. X Ges.
Math. U'erke, i. pp. 67, 75.



3-22-3'3l] CONTINUOUS FUNCTIONS AND UNIFORM CONVERGENCE 45

Let the functions u (z), Wg (z), ... be defined at all points of a
closed region of the Argand diagram. Let

Sn (z) = U, (z) + M2 (2) + ...+ tin Z). '

oc

The condition that the series 2 Un z) should converge for any
particular

n = l

value of z is that, given e, a number n should exist such that

I n+p z) — Sn (z)\ < €

for all positive values of jo, the value of 7i of course depending on
e.

Let n have the smallest integer value for which the condition is
satisfied. This integer will in general depend on the particular value
of z which has been selected for consideration. We denote this
dependence by writing n (z) in place of ?i. Now it mag happen that we
can find a number N,

INDEPENDENT OF Z, SUch that

n z)<N "' '"

for all values of z in the region under consideration.

If this number N exists, the series is said to converge uniformly
throughout the region.

If no such number iV exists, the convergence is said to be
non-uniform*.

Uniformity of convergence is thus a property depending on a whole set
of values of z, whereas previously we have considered the convergence
of a series for various particular values of z, the convergence for
each value being con- sidered without reference to the other values.

We define the phrase ' uniformity of convergence near a point z ' to
mean that there is a definite positive number 8 such that the series
converges uniformly in the domain common to the circle \ z — z \ \ h
and the region in which the series converges.

3'31. On the condition for uniformity of convergence' .

If Rn,p z) = Un+i z) + iin+2 (z) + ... + Un+p (z), WO have Seen that
the

necessary and sufficient condition that S Un (z) should converge
uniformly

in a region is that, given any positive number e, it should be
possible to choose N INDEPENDENT OF z (but depending on e) such that

I Rn, p(z)\ < €

for ALL positive integral values of p.

* The reader who is unacquainted with the concept of uniformity of
convergence will find it made much clearer by consultinf; Bromwich,
Iiijinite Series, Ch. vii, where an illuminating account of Osgood's
graphical investigation is given.

t This section shews that it is indifferent whether uniformity of
convergence is defined by means of the partial remainder Rj p(z) or by
iJ (2). Writers differ in the definition taken as fundamental.



46 THE PROCESSES OF ANALYSIS [CHAP. Ill

If the condition is satisfied, by § 2-22, 8n z) tends to a limit, S
z), say for each value of under consideration; and then, since e is
independent of p,

and therefore, when n > iV,

S (2) - Sn (Z) = I lim i v, p 2) - E v. n-N (z), /

and so \ S(z)-S iz)\ < 2€.

Thus (writing e for e) a necessary condition for uniformity of
convergence is that \ S z) — Sn (z) \ < e, whenever n>N and N is
independent of z; the condition is also sufficient; for if it is
satisfied it follows as in § 2-22 (I) that I R rp z) 1 < 2e, which, by
definition, is the condition for uniformity.

Example 1. Shew that, if x be real, the sum of the series

X X X

TODO

is discontinuous at .r=0 and the series is non-uniformly convergent
near ' = 0.

The sum of the first n terms is easily seen to be 1; so when x — Q
the

nx+i

sum is; when a* 4=0, the sum is 1.

1 ' The value of RnXx) = S x)-Sn x) is — -- if x O; so when x is
small, say

:r=one-hundred-milliouth, the remainder after a million terms is — or
l-TTyTj

100 + 1 the first million terms of the series do not contribute one
per cent, of the sum. And in

general, to make < e, it is necessary to take

° nx + 1

7i>-(--l

X \ e

Corresponding to a given e, no number N exists, independent of x, such
that n<N for all values of x in any interval including x =; for by
taking x sufficiently small we can make n greater than any number N
which is independent of x. There is therefore non- uniform convergence
near ,r = 0.

Example 2. Discuss the series

x n. n±V)x'''-\ ]



nZx + n x \ + 7i- \ fx y in which x is real.

r. , , , . nx (n- ) X , . X ,

The 7ith term can be written , 5—5 - , —; , . — q , so o (x) = n ,
and

l+n\ v l + n + \ y X'' l+x''

  W ] +(,i + 1)2 2 •

[Note. In this example the sum of the series is not discontinuous at =
0.] But (taking 6<i, and 4=0), \ Rn x)\ < e if e-i(?i + l) \ x\ < l-ir
n+\ f x; i.e. if ?i+l>| €-i + Ve-2-4 |.T|-l or ?t + l<J e-i-v'e = 3-4
|:p|- .



3*32] CONTINUOUS FUNCTIONS AND UNIFORM CONVERGENCE 47

Now it is not the case that the second inequality is satisfied for all
values of n greater than a certain value and for all values of x; and
the first inequality gives a value of n(x) which tends to infinity as
x- 0; so that, corresponding to any interval containing the point =0,
there is no number N' independent of x. The series, therefore, is
non-uniformly convergent near :i, = 0.

The reader will observe that n x) is discontinuous at .r = 0; for n
x)- 'x> as .r- -0, but n(0) = 0.

3'32. Connexion of discontinuity luith non-uniform convergence.

We shall now shew that if a semes of continuous functions of z is
uniformly convergent for all values of z in a given closed domain, the
sum is a continuous function of z at all points of the domain.

For let the series be f z) = it, z) + Wo (2 ) + • . . + Un z)+ ...— Sn
(z) + Rn z), where Rn (z) is the remainder after n terms.

Since the series is uniformly convergent, given any positive number e,
we

can find a corresponding integer n independent of z, such that | R ( )
| < e

for all values of z within the domain.

Now n and e being thus fixed, we can, on account of the continuity of
Sn (•2 ), find a positive number rj such that

\ Sn(z)-S z')\ < l€,

whenever \ z — z' \ < r . We have then

l/( ) -/V) I = 1 [Sni ) - Sn z')] \ + lRn(2)- Rn ) \ < ! Sn z) -
Sn(z') I + I Rniz) \ + | Rn z') |

' < 6,

which is the condition for continuity at z. Example 1. Shew that near
x = the series

Ui x) + U<i x) + M3 ( 0 + • • • ,

1 1

where i(vP)=a;, (a;) =;p "~* — :p "~ ,

and real values of x are concerned, is discontinuous and non-uniformly
convergent.

In this example it is convenient to take a slightly different form of
the test; we shall shew that, given an arbitrarily small number f, it
is possible to choose values of x, as small as we please, depending on
n in such a way that | R [x) \ is not less than e for any value of n,
no matter how large. The reader will easily see that the existence of
such values of x is inconsistent with the condition for uniformity of
convergence. i\

The value of S ix) is .r" "-i; as n tends to infinity, S x) tends to
1, 0, or - 1, accord- ing as x is positive, zero, or negative. The
series is therefore absolutely convergent for all values of or, and
has a discontinuity at .t' = 0.



48 THE PROCESSES OF ANALYSIS [cHAP. Ill

1

In this series R x) = l-x ''-\ x > 0); however great n may be, by
taking* x = e- (2" - 1) we can cause this remainder to take the value
l-e' which is not arbitrarily small. The series is therefore
non-uniformly convergent near .r = 0.

Example 2. Shew that near z = the series

 il H-(l+2) -l l+(l + 2)"

is non-uniformly convergent and its sum is discontinuous. The nth term
can be written

l-(l+3)" 1-(1+Z)"-

l + (H-3) H-(l+z) -i' so the sum of the first n terms is , — — . Thus,
considering real values of z greater

than - 1, it is seen that the sum to infinity is 1, 0, or — 1,
according as z is negative, zero,

or positive. There is thus a discontinuity at 2 = 0. This
discontinuity is explained by the

fact that the series is non-uniformly convergent near 2=0; for the
remainder after n terms

in the series when z is positive is

-2

r+rr+2)"'

and, however great n may be, by taking z = ~, this can be made
numerically greater

2

than - , which is not arbitrarilv small. The series is therefore
non-uniformlv con- 1-1-e'

vergent near 2 = 0.

3'33. The distinction between absolute and uyiiform convergence.

The uniform convergence of a series in a domain does not necessitate
its absolute convergence at any points of the domain, nor conversely.
Thus

the series S vz r converges absolutely, but (near z = 0) not uniformly
;

(1 + ")"

while in the case of the series

 =i z + n ' the series of moduli is

1

5", =



 =i \ n + Z'\ \ which is divergent, so the series is only
conditionally convergent; but for all real values of z, the terms of
the series are alternately positive and negative and numerically
decreasing, so the sum of the series lies between the sum of its first
n terms and of its first (n -f 1) terms, and so the remainder after n
terms is numerically less than the nth. term. Thus we only need take a
finite number (independent of z) of terms in order to ensure that for
all real values of z the remainder is less than any assigned number e,
and so the series is uniformly convergent.

Absolutely convergent series behave like series with a finite number
of terms in that we can multiply them together and transpose their
terms.

* This value of x satisfies the condition i a; | < 5 whenever 2rt - 1
> log 5~i.



3'33-3"341] CONTINUOUS functions and uniform convergence 49

Uniformly convergent series behave like series with a finite number of
terms in that they are continuous if each term in the series is
continuous and (as we shall see) the series can then be integrated
term by term.

334. A condition, due to Weierstrass* , for uniform convergence.

A sufficient, though not necessary, condition for the uniform
convergence of a series may be enunciated as follows : —

If, for all values of z within a domain, the moduli of the terms of a
series *S' = Ui (z) + i<2 ( ) + W3 ( ) + • • • are respectively less
than the corresponding terms in a convergent series of positive terms

where M is independent of z, then the series S is uniformly convergent
in this region. This follows from the fact that, the series T being
convergent, it is always possible to choose n so that the remainder
after the first n terms of T, and therefore the modulus of the
remainder after the first n terms of *Si, is less than an assigned
positive number e; and since the value of n thus found is independent
of z, it follows (§ 3-31) that the series S is uni- formly convergent
; by § 234, the series S also converges absolutely.

Example. The .scries

1 ., 1 ,

cos Z + , CO.S- 2+57, CO.S-* Z- ...

is uniformly convergent for all re<il values of r, because the moduli
of its terms are not greater than the corresi)onding terms of the
convergent series

I i whose terms are positive constants.

3 •341 . Uniformity of convergence of infinite products t.

A convergent product n 1 + ?< (z) is said to converge uniformly in a
domain of values

M=l

of z if, given e, we can tind m independent of z such that

n 1 + (z) - n i+u z) \ < €

for all positive integral values of p.

The only condition for imiformity of convergence which will be used in
this work is that the product converges uniformly if | m (s) | < J/
where J/ is independent of 2 and

2 J/ convei-ges.

n = l

* Abhandlungen aus der Funktionenlehre, p. 70. The test given by this
condition is usually described (e.g. by Osgood, Annals of Mathematics,
iii. j[1889), p. 130) as the M-test.

t The definition is, effectively, that given by Osgood,
Funktionentheorie, p. 462. The condition here given for uuiformity of
convergence is also established in that work.

W. M. A. 4



50 THE PROCESSES OF ANALYSIS [CHAP. Ill

To prove the validity of the condition we observe that n (l + J/ )
converges (§ 2-7),

M = l

and so we can choose m such that

Vl+p VI

n l + M,, - n l + M <€;

n = l )( = 1

and then we have

m+p m I I m p ni+p ~] I

n l+un (.-) - n 1 +t< (s) =1 n 1 + (z) n i +u (z) - 1

11=1 M=l I I n=l \ \ n=m+i J I

m r ni+2> ~]

 n(i + 14) n i+i/ -i

m=l L n = m+l J

and the choice of ?h is independent of z.

3 "35. Hardy's tests for uniform convergence*.



The reader will see, from § 2-31, that if, in a given domain.



p



2 a z) k where a (2) is



real and k is finite and independent of and 2, and if / (2) t + i (s)
and fn )-

uniformly as w - - oo , then 2 a z) f £) converges uniformly. t=i

Also that if where k is independent of 2 and 2 a (2) converges
uniformly, then 2 a,j £) w (s) con-

Ii = l M = l

verges uniformly. [To prove the latter, observe that m can be found
such that

 7n + l(2), m + l(2)+ 7 + 2(2), •••, m + 1 ( ) + m + 2 ( )+ • • + m +
p (2)

are numerically less than e\ k; and therefore (§ 2-301)

2 a (2) M (s) < e <,n+i (2)//(-< e, n=jn+l I

and the choice of e and hi is independent of 2.]



° cos nQ "" sin % 2 , 2

7t=l '* n = l 'i



Example 1. Shew that, if S>0, the series

converge uniformly in the range

S (9 27r - S.

Obtain the corresponding result for the series

  (-)"cos?i<9 ' ( - ) sin n6

2i , 2 ,

n=l n n=\ n

by writing O + n for .

Example 2. If, when a a,' 6, | co,i (.1;) | < -j and 2 | < + i (.r) —
co (.r) | <j('2, where

)(=i

 •1, k.> are independent of n and .r, and if 2 a is a convergent
series independent of x,

n=i

then 2 a,tC > (:*;) converges uniformly when a .r / . (Hardy.)

n = l

* Proc. London Math. Soc. (2) iv. (1907), pp. 247-265. These results,
which are generalisa- tions of Abel's theorem (§ 3-71, below), though
well known, do not appear to have been published before 1907. From
their resemblance to the tests of Dirichlet and Abel for convergence,
Bromwich proposes to call them Dirichlet's and Abel's tests
respectively.



3"35, 3"4] CONTINUOUS FUNCTIONS AND UNIFORM CONVERGENCE 51

[For we can choose m, independent of .v, such that corollary, we have



m+p I

2 a < e, and then, l)y § 2-301

n=m+l I



m+p I

2 ttnOin ( •) I < ( 'l + ' '2) f •]



n=m+l I

3 "4. Discussion of a particular double senes.

Let (1)1 and w. be any constants whose ratio is not purely real; and
let a be positive.

The series 2; r- , in which the summation extends over

all positive and negative integral and zero values of in and n, is of
great importance in the theory of Elliptic Functions. At each of the
points z = — 2mcoi — 2/10)2 the series does not exist. It can be shewn
that the series converges absolutely for all other values of if a > 2,
and the convergence is uniform for those values of z such that; 2 +
2niQ)i + 2nco2 8 for all integral values of m and n, where 8 is an
arbitrary positive number.

Let S' denote a summation for all integral values of m and n, the term
for which 7n = n = being omitted.

Now, if ni and n are not both zero, and if \ z + 2m(o + 2nco.2\ " 8 >
for all integral values of m and n, then we can find a positive number
C. de- pending on B but not on 2, such that

I 1 !



(z + 2w\&)i + 2 a)a)''



 2m(Oi + 2710)2)"



Consequently, by § 3"34, the given series is absolutely and uniformly*
convergent in the domain considered if

2' 1

I mcoi + no)2 1 " converges.

To discuss the convergence of the latter series, let

0), = CTj 4- 1/3, , 0)0 = Qfo + z'/So ,

where a , a.,, /3i, /Sa are real. Since co../ coi is not real, a /S. —
ou/3i 4= 0. Then the series is

2'

 (a,m + aojiy- + /3,m + /SoTi) *

This converges (§ 2"5 corollary) if the series

S = ' i — -

(m-+n2)

converges; for the quotient of corresponding terms is

The reader will easily define uniformity of convergence of double
series (see § 3-5).

4—2



52 THE PROCESSES OF ANALYSIS [cHAP. Ill

where /z = njm. This expression, qua function of a continuous real
variable jx, can be proved to have a positive minimum* (not zero)
since ofi/3o — ao/3i =|=; and so the quotient is always greater than
a positive number K (independent of/x).

We have therefore only to study the convergence of the series S. Let

PI 1

V V S'



 'p,q



00 00 1

 4 S S' - , .

m = n=0 (m + 71 )2 *

Separating Sp g into the terms for which m = n, m > n, and m < n, re-
spectively, we have

pi p m-l 1 q n-1 1

IS. = S - I- s S - + s s .

'% 1 ml

But S r- <



n=o ni" + n ) "- (m ) "



m°



Therefore IS t -J— + S — + i .

But these last series are known to be convergent if a — 1 > 1. So the
series S is convergent if a > 2. The original series is therefore
absolutely and uni- formly convergent, when a > 2, for the specified
range of values of z.

Example. Prove that the series

1

2



(mj + 7112 + . . . + my?y

in which the summation extends over all positive and negative integral
values and zero values of mj, m2, ... wi , except the set of
simultaneous zero values, is absolutely convergent if fi>ir.
(Eisenstein, Journal fur Math, xxxv.)

3'5. The concept of uniformity.

There are processes other than that of summing a series in which the
idea of uniformity is of importance.

Let e be an arbitrary positive number; and let f z, ) be a function of
two variables z and , which, for each point z oi a, closed region,
satisfies the inequality \ f z, ) | < 6 when t, is given any one of a
certain set of values which will be denoted by ( z); the particular
set of values of course depends on the particular value of z under
consideration. If a set ( )o can be found such that every member of
the set ( )o is a member of all the sets ( j), the function f z, ) is
said to satisfy the inequality uniformly for all points z of

* The reader will find no difficulty in verifying this statement; the
minimum value in question is given by

K "' = h W + a + . + r- (a,-/3,r-+(a2 + ,3i)=l (ai + /32)2+ (a - iP ]-



3"5, 3'6] CONTINUOUS FUNCTIONS AND UNIFORM CONVERGENCE 53

the region. And if a function (f> z) possesses some property, for
every positive value of e, in virtue of the inequality \ f z, ):<e,(f)
(z) is then said to possess the property uniformly.

In addition to the uniformity of convergence of series and products,
we shall have to consider uniformity of convergence of integrals and
also uniformity of continuity; thus a series is uniformly convergent
when \ R, z)\ <e, t( = 0 assuming integer values in- dependent of z
only.

Further, a function f z) is continuous in a closed region if, given e,
we can find a I ositive number r/ such that 1/(2 + 2) —fi ) \ <
whenever

0<\ C \ < r and 2 + f is a point of the region.

The function will be uniformly continuous if we can find a positive
number; inde- pendent of z, such that rjKr and \ f z + C)~f ) i <
whenever

0<UI<'7 and 2 + f is a point of the region, (in this case the set (f)o
is the set of points whose moduli are less than r)).

We shall find later (§ 3-61) that continuity involves uniformity of
continuity; this is in marked contradistinction to the fact that
convergence does not involve uniformity of convergence.

36. The modified Heine-Borel theorem.

The following theorem is of great importance in connexion with
properties of uniformity; we give a proof for a one-dimensional
closed region*.

Given (i) a straight line CD and (ii) a latv by which, corresponding
to each point f P of CD, we can determine a closed interval I P) of
CD, P being an interior point of I (P).

Tlien the line CD can be divided into a finite number of closed
intervals Ji, Jo, ... Jk, such that each interval Jr contains at least
one point not an end point) Pr, such that no point of Jr lies outside
the interval I (Pr) associated (by means of the given law) luitli that
point Pr§.

A closed interval of the nature just described will be called a
suitable interval, and will be said to satisfy condition A).

If CD satisfies condition A ), what is required is proved. If not,
bisect CD; if either or both of the intervals into which CD is
divided is not- suitable, bisect it or them||.

* A formal proof of the tlieorem for a two-dimensional region will be
found in Watson's Complex Integration and Cauchy s Theorem (Camb.
Math. Tracts, No. 15).

t Examples of such laws associating intervals with points will be
found in §§ 3'61, 5'13.

t Except when P is at C or D, when it is an end point.

§ This statement of the Heine-Borel theorem (which is sometimes called
the Borel-Lebesgue theorem) is due to Baker, Proc. London Math. Soc.
(2) i. (1904), p. 24. Hobson, The Theonj of Functions of a Real
Variable (1907), p. 87, points out that the theorem is practically
given in Goursat's proof of Cauchy's theorem Trans. American Math.
Soc. i. (1900), j). 14); the ordinary form of the Heine-Borel theorem
will be found in the treatise cited.

II A suitable interval is not to be bisected; for one of the parts
into which it is divided might not be suitable.



54 THE PROCESSES OF ANALYSIS [CHAP. Ill

This process of bisecting intervals which are not suitable either will
terminate or it will not. If it does terminate, the theorem is proved,
for CD will have been divided into suitable intervals.

Suppose that the process does not terminate; and let an interval,
which can be divided into suitable intervals by the process of
bisection just described, be said to satisfy condition (B).

Then, by hypothesis, CD does not satisfy condition B); therefore at
least one of the bisected portions of CD does not satisfy condition
B). Take that one which does not (if neither satisfies condition B)
take the left-hand one); bisect it and select that bisected part
which does not satisfy condition B). This process of bisection and
selection gives an unending sequence of intervals 5o, Si, S2, ... such
that :

(i) The length of s is 2-" Ci).

(ii) No point of s,i+i is outside Sn-

(iii) The interval s does not satisfy condition (-4).

Let the distances of the end points of s from G be Xn, yn\ then Xn <
a? +i < 2 i+i Hn- Therefore, by § 2*2, x and yn have limits; and, by
the condition (i) above, these limits are the same, say; let Q be the
point whose distance from C is . But, by hypothesis, there is a number
hq such that every point of CD, whose distance from Q is less than 5q,
is a point of the associated interval /(Q). Choose n so large that t
CDk 8q; then Q is an internal point or end point of Sn and the
distance of every point of Sn from Q is less than Sq. And therefore
the interval 5 satisfies condition (A), which is contrary to condition
(iii) above. The hypothesis that the process of bisecting intervals
does not terminate therefore involves a contradiction; therefore the
process does terminate and the theorem is proved.

In the two-dimensional form of the theorem* the interval CD is
replaced by a closed two-dimensional region, the interval I P) by a
circlet with centre P, and the interval Jj. by a square with sides
parallel to the axes.

3'61. Uniformity of continuity.

From the theorem just proved, it follows without difficulty that if a
function f(x) of a real variable x is continuous when a x b, then f(x)
is unifurmly continuous throughout the range a x b.

For let e be an arbitrary positive number; then, in virtue of the
con- tinuity of f x), corresponding to any value of x, we can find a
positive number S , depending on x, such that

1/( 0 -/( O I < e

for all values of x' such that \ x' — x\ < Sx-

* The reader will see that a proof may be constructed on similar lines
by drawing a square circumscribing the region and carrying out a
process of dividing squares into four equal squares.

t Or the portion of the circle which lies inside the region.

:J: This result is due to Heine; see Journal fiir Math. lxxi. (1870),
p. 361, and lxxiv. (1872), p. 188.



3'61, 3'62] CONTINUOUS functions and uniform convergence 55

Then by § 3"G we can divide the range (a, b) into a finite number of
closed intervals with the property that in each interval there is a
number Xi such

that \ fioc') — f xx) \ < -€, whenever x lies in the interval in which
a;, lies.

Let So be the length of the smallest of these intervals; and let f ,
|' be any two numbers in the closed range (a, 6) such that | — ' | <
o- Then , f ' lie in the same or in adjacent intervals; if they lie
in adjacent intervals let , be the common end point. Then we can find
numbers x , Xo, one in each interval, such that

\ f )-f .)\ < \ e, ,/(?o)-/(- 0 < f>

' /(r ) -/( e) \ < \ \ , fit) -fu i < 5 6,

so that

i/(i) -/(r) = , /(B -fM] - [fit) -fM]

- /(r)-/( -.) + /( o)-/( .) i

< 6.

If , ' lie in the same interval, we can prove similarly that

i/( )-/(r)i<2 -

In either case we have shewn that, for (iny number in the range, we
have

\ f( )-f +0 <e whenever + is in the range and —Bo< < Bq, where So is
independent of . The uniformity of the continuity is therefore
established.

Corollary (i). From the two-dimensional form of the theorem of § 3"6
we can prove that a function of a complex variable, continuous at all
jwints of a closed region of the Argand diagram, is uniformly
continuous throughout that region.

Corollary (ii). A function f x) which is continuous throughout the
range a x b is hounded in the range; that is to say we can find a
number k independent of x such that \ f x) ! <K for all points x in
the range.

[Let n be the number of parts into which the range is divided.

Let , i, 2> ••• In-ij be their end points ] then if x be any point of
the rth interval we can find numbers Xi, x-i, ... Xn such that

l/( )-/(- 'i)|< , l/(.i-i)-/'(li)|<i , l/( i)-/(' 2)|<ie, l/(. 2)-/(
2)|<if,...

••• \ f ryi)-f x)\ < h. Therefore \ f a)-f x) |< ire, and so

which is the required result, since the right-hand side is independent
of x."]

The corresponding theorem for functions of complex variables is left
to the reader.

3'62. A real function, of a real variable, continuous in a closed
interval, attains its upper bound.

Let f x) be a real continuous function of x when a x b. Form a section
in which the i?-class consists of those numbers r such that r >f x)



56 THE PROCESSES OF ANALYSIS [cHAP. Ill

for all values of x in the range (a, h), and the X-class of all other
numbers. This section defines a number a such that f x) a., but, if h
be any positive number, values of x in the range exist such that
f(x)>a — 8. Then a is called the upper bound oi f x); and the theorem
states that a number x' in the range can be found such thai f x) = a.

For, no matter how small h may be, we can find values of x for which
|/( ) — aj" >'8~; therefore | /(. ) - a| |~ is not bounded in the
range; therefore (§ 3'61 cor. (ii)) it is not continuous at some point
or points of the range; but since | f x) — a | is continuous at all
points of the range, its re- ciprocal is continuous at all points of
the range (§ 3*2 example) except those points at which f(x) = a',
therefore f x) = a at some point of the range; the theorem is
therefore proved.

Corollary (i). The lower bound of a continuous function may be defined
in a similar manner; and a continuous function attains its lower
bound.

Corollary (ii). If /( ) be a function of a complex variable continuous
in a closed region, | f z) \ attains its upper bound.

3'63. A real function, of a. real variable, continuous in a closed
interval, attains all values between its upper and loiuer bounds.

Let 31, m be the upper and lower bounds off x); then we can find
numbers X, -v, by § 362, such that/( ) = M,f x) = m; let //. be any
number such that m< fjb< M. Given any positive number e, we can (by §
3-61) divide the range (x, x) into 'A finite number, r, of closed
intervals such that

l/(.r,'-')-/(*2' ')|<6,

where a i""*, iCjC* are any points of the rth interval; take -i**"',
x./'' to be the end points of the interval; then there is at least
one of the intervals for which /(*'!<' ') - f ,f(x.J ) — /x have
opposite signs; and since

|[/( ,<'-')-/z - /( ,"-))- j|<6,

it follows that j /(aa""') — /ii\ < e.

Since we can find a number i'''* to satisfy this inequality for all
values of 6, no matter how small, the lower bound of the function \
f(x)-fA,\ is zero; since this is a continuous function of x, it
follows from § 3-62 cor. (i) that/( ) — yLt vanishes for some value of
. .

3'64. The fluctuation of a function of a real variable*.

Let/(;) be a real bounded function, defined when a x b. Let

a Xi X2 ... : Xn b. Then I /(a) -f(x,) \ + \ f(x,) -f(x,) | + ... +
l/C J -f(b) | is called the fluctuation oi f x) in the range (a, b)
for the set of subdivisions x , X2, ... Xn.

The terminology of tliis section is partly that of Hobson, The Theory
of Functions of a Real Variable (1907) and partly that of Young, The
Theory of Sets of Points (190(5).



3'63-37l] CONTINUOUS FUNCTIONS AND UNIFORM CONVERGENCE 57

If the fluctuation have an upper bound FJ*, independent of n, for all
choices of iCi, j-o, ... Xn, then f x) is said to have limited total
fluctuation in the range (a, h). Fa!' is called the total fluctuation
in the range.

Example 1. If f x) be monotonic* in the range (a, b), its total
fluctnation in the range is|/(a)-/(6)|. '>i

Example 2. A fimction with limited total fluctuation can be expressed
as the differ- ence of two positive increasing monotonic functions.

[These function.s may be taken to be | Fa' +fix) , h Fa'-f x) .]

Example 3. If f x) have limited total fluctuation in the range a, b),
then the limits f x±0) exist at all points in the interior of the
range. [See § 3*2 example.]

Example 4. li f x\ g x) have limited total fluctuation in the range
(a, b) so has f x)g x).

[For \ f x')g x')-f x)g x)\ \ \ f :>f). \ g x')-g x)\ + \ g ) \ f
')-fi. )l and so the total fluctuation of f x) g x) cannot exceed g .
FJ'+f. G K where g are the upper bounds of |/( ) |, \ g (x) \ .]

- 3'7. Uniformity of convergence of power series. Let the power
.series

ao + tti;-!- ... +an2"+ ••• '

converge absolutely when z = Zo.

Then, if | | | 'o | , , a " | | a Zo" | .

00 CO

But since S | anZj" ' converges, it follows, by § 3-34, that S ttn "
converges

w = ' 71 =

uniformly with regard to the variable z when \ z \ Zq.

Hence, by § 3"32, a power series is a continuous function of the
variable throughout the closed region formed by the interior and
boundary of any circle concentric with the circle of convergence and
of smaller radius (§ 2"6).

  3'71. Abel's theoreni-'r on continuity up to the circle of
convergence.

00

Let S a,j " be a po ver series, whose radius of convergence is unity,
and

M =

cc

let it be such that S a converges; and let a; 1; then Abel's theorem

>i =

(OC \ 00

S '*] = S an. . n=0 J M=0

For, with the notation of § 3-35, the function x satisfies the
conditions

00

laid on u, x), when 0 a; l; consequently /(a:;) = 2 On " converges
uni-

M =

* The function is monotonic if f x)-f x')\ \ x-x') is one-signed or
zero for all pairs of different values of .r and x' .

t Journal fiir Math. i. (1826), pp. 311-339, Theorem iv. Abel's proof
employs directlj' the arguments by which the theorems of § 3-32 and §
3-35 are proved. In the case when S | a I converges, the theorem is
obvious from § 3-7.



58 THE PROCESSES OF ANALYSIS [CHAP. Ill

formly throughout the range a;: 1; it is therefore, by § 3"32, a
continuous function of x throughout the range, and so lim f x)=f l),
which is the

a;-*l-0

theorem stated.

3'72. Abel's theorem* on multiplication of series.

This is a modification of the theorem of § 2"53 for absolutely
convergent series.

Let Cn = ttohn + i \& \ i + . . . + an 0 •

Then the convergence of " an, S bn and 2 c is a sufficient condition
that

M=0 n=0 w=0

 00 \ / 00 \ 00

For, let ,

 (a;)= 2 an B(x)= 2 6na ", C(a.')-= i CnX''. n=0 n=0 M=0

Then the series for A (x), B x), C(x) are absolutely convergent when I
I < 1, (§ 2-6); and consequently, by § 2-53,

A(x)B(x)=C x) when < < 1; therefore, by § 2-2 example 2,

  lim A x)] lim B(x) = lim C(x)\

a; .l-0 a;- -l-0 x -l-O

provided that these three limits exist; but, by § 3-71, these three
limits are

00 00 30

2 a , 'Z bn, % Cn', and the theorem is proved.

n=0 w = w=0

3'73. Power series which vanish identically.

If a convergent poiver seynes vanishes for all values of z such that \
z\ r- , where r > 0, then all the coefficients in the power series
vanish.

For, if not, let a be the first coefficient which does not vanish.

Then am + '7/t+i + m+2' "+ ... vanishes for all values of z (zero
excepted) and converges absolutely when \ z\ \ r<r; hence, if s = a
+i + a,n+2Z + . . ., we have

00

I S I Z, I (ljn-|-7i I T , >j = l

and so we can findf a positive number S r such that, whenever | | S,

I , 2 \ L I 1 ' I

I ( m+i + ( f"m+2 + • • • 1 2' '" I '

and then | a + s | | m | - ! •' ' > i a.m\, and so a, + s =f when \ z
\ < S.

* Journal filr Math. i. (1826), pp.;:ili-339, Theorem vi. This is
Abel's original proof. In some text-books a more elaborate proof, by
the use of Cesaro's sums (§ 8-43), is given.



372,373] CONTINUOUS FUNCTIONS AND UNIFORM CONVERGENCE 59

We have therefore arrived at a contradiction by supposing that some
coefficient does not vanish. Therefore all the coefficients vanish.

Corollary 1. "We may 'equate corresponding coefficients' in two power
series whose sums are equal throughout the region 1 |<S, where 8>0.

Corollary 2. We may also equate coefficients in two power series which
are proved equal only when z is real.



REFERENCES.

T. J. 1'a. Bromwich, Theory of Infinite Series (1908), Ch. vii.

E. GouRSAT, Cours d'Analyse (Paris, 1910, 1911), Chs. i, xiv.

C J. DE LA Vall e Poussin (Louvain and Paris, 1914), Cours d Analyse
Infinitesimale, Introduction and Ch. viii.

G. H. Hardy, A course of Pure Mathematics (1914), Ch. v.

VV. F. Osgood, Lehrbiich der Funktionentheorie (Leipzig, 1912), Chs.
ii, iii.

G. N. Watson, Complex Integration and Cauchy's Theorem (Camb. Math.
Tracts, No. 15), (1914), Chs. I, II.



Miscellaneous Examples.

1. Shew that the series



 =l(l-2 )(l-2 l)

is equal to ,-, —. when 1 2 I < 1 and is equal to ,, -, when 1 2 1 >
1.

  zy z z)-

Is this fact connected with the theory of uniform convergence ?

2. Shew that the series

2sini + 4sinl + ... + 2"sin + ...

converges absolutely for all values of z z = excepted), but does not
converge uniformly near 2=0.

3. If Un x)=-2 n-\ f .re-'"-''' ' + n xe''' '''',

shew that 2 ?< x) does not converge uniformly near x=0. (Math. Trip.,
1907.)

n = \

4. Shew that the series —pr j- + —r —... is convergent, but that its
square (formed

by Abel's rule)

T 2+1,73 + 2; U'4' vW -

is divergent.

5. If the convergent series 5=— — — +—-—+... (r>0) be multiplied by
itsel,

the teiTus of the product being arranged as in Abel's result, shew
that the resulting series diverges if r | but converges to the sum s \
i r>. (Cauchy and Cajori.)



60 THE PROCESSES OF ANALYSIS [CHAP. Ill

6. If the two conditionally convergent series

2 - and 2



n=i W =i n'

where r and s lie between and 1, be multiplied together, and the
product an-anged as in Abel's result, shew that the necessary and
sufficient condition for the convergence of the resulting series is r
+ s >1. (Cajori.)

7. Shew that if the series 1 - 3 + 5 - t + • • •

be multiplied by itself any number of times, the terms of the product
being arranged as in Abel's result, the resulting series converges.
(Cajori.)

8. Shew that the qth. power of the series

 ! sin d+a sin 2 + ... +a,i sin n6 + ... is convergent whenever 5- (1
- r)< 1, r being the greatest number satisfying the relation

for all values of n.

9. Shew that if 6 is not equal to or a multiple of 2n, and if %0) Ui,
u , ...he a sequence such that u - O steadily, then the series 2i/ cos
nd + a) is convergent.

Shew also that, if the limit of t< is not zero, but ?i is still
monotonic, the sum of the

S

series is oscillatory if - is rational, but that, if - is irrational,
the sum may have any value

TT TT

between certain bounds whose difference is a cosec 6, where a= lim u -

(Math. Trip., 1896.)

