\chapter{The Equations of Mathematical Physics} 

18"1. The differential equations of mathematical pliysics. 

The functions which have been introduced in the preceding chapters are 
of importance in the applications of mathematics to physical investigations. 
Such applications are outside the province of this book ; but most of them 
depend essentially on the fact that, by means of these functions, it is possible 
to construct solutions of certain partial differential equations, of which the 
following are among the most important : 

(I) Laplace s equation 

dx- dy'  dz  
which was originally introduced in a memoir* on Saturn's rings. 

If  x, y, z) be the rectangular coordinates of any point in space, this equation is 
satisfied by the following functions which occur in various branches of mathematical 
physics : 

(i) The gravitational potential in regions not occupied by attractii:ig matter. 

(ii) The electrostatic potential in a uniform dielectric, in the theory of electro- 
statics. 

(iii) The magnetic potential in free aether, in the theory of magnetostatics. 

(iv) The electric potential, in the theory of the steady flow of electric currents in 
solid conductoi's. 

(v) The temperature, in the theory of thermal equilibrium in solids. 

(vi) The velocity potential at points of a homogeneous liquid moving irrotationally, 
in hydrodynamical problems, 

Notwithstanding the physical diflferences of these theories, the mathematical investi- 
gations are much the same for all of them : thus, the problem of thermal equilibrium in a 
solid when the points of its surface are maintained at given temperatures is mathe- 
matically iden.tical with the problem of determining the electric intensity in a region 
when the points of its boundary are maintained at given potentials. 

(II) The equation of ivave motions 

dx  dy'  dz  C' dt- 
This equation is of general occurrence in investigations of undidatory disturbances 
propagated with velocity c independent of the wave length ; for example, in the theory of 
electric waves and the electro-magnetic theory of light, it is the equation satisfied by each 
component of the electric or magnetic vector; in the theory of elastic vibrations, it 
is the equati<Mi satisfied by each component of the displacement ; and in the theory 
of sound, it is the equation satisfied by the velocity potential in a perfect gas. 

* Mem. de FAcad. des Sciences, 1787 (published 1789), p. 252. 



18*1, 18 "2] THE EQUATIONS OF MATHEMATICAL PHYSICS 387 

(III) The equation of conduction of heat 

dfV dfV dfV\ ldV 
dec- dy  dz- k dt 

This is the equation satisfied by the temperature at a point of a homogeneous isotropic 
body ; the constant k is proportional to the heat conductivity of the body and inversely 
proportional to its specific heat and density. 

(IV) A particular case of the preceding equation (II), when the variable 
z is absent, is 

dx- dy- c- dt'  

This is the equation satisfied by the displacement in the theory of transverse vibrations 
of a membrane ; the equation also occurs in the theory of wave motion in two dimensions. 

(V) The equation of telegraphy 

This is the equation satisfied by the potential in a telegraph cable when the inductance 
X, the capacity K, and the resistance It per unit length are taken into account. 

It would not be possible, within the limits of this chapter, to attempt 
an exhaustive account of the theories of these and the other differential 
equations of mathematical physics; but, by considering selected 'typical 
cases, we shall expound some of the principal methods employed, with 
special reference to the uses of the transcendental functions. 

18'2. Boundary conditions. 

A problem which arises very frequently is the determination, for one of the 
 equations of § 18*1, of a solution which is subject to certain boundary con- 
ditions ; thus we may desire to find the temperature at any point inside a 
homogeneous isotropic conducting solid in thermal equilibrium when the 
points of its outer surface are maintained at given temperatures. This 
amounts to finding a solution of Laplace's equation at points inside a given 
surface, when the value of the solution at points on the surface is given. 

A more complicated problem of a similar nature occurs in discussing 
small oscillations of a liquid in a basin, the liquid being exposed to the 
atmosphere ; in this problem we are given, effectively, the velocity potential 
at points of the free surface and the normal derivate of the velocity potential 
where the liquid is in contact with the basin. 

The nature of the boundary conditions, necessary to determine a solution 
uniquely, varies very much with the form of differential equation considered, 
even in the case of equations which, at first sight, seem very much alike. 
Thus a solution of the equation 



dx- dy'  



25—2 



388 THE TRANSCENDENTAL FUNCTIONS [CHAP. XVIII 

(which occurs in the problem of thermal equilibrium in a conducting 
cylinder) is uniquely determined at points inside a closed curve in the 
x7/-ip\ a,ne by a knowledge of the value of V at points on the curve ; but 
in the case of the equation 

da;  c- dt- 

(which effectively only differs from the former in a change of sign), occurring 

in connexion with transverse vibrations of a stretched string, where V 

denotes the displacement at time t at distance x from the end of the 

string, it is physically evident that a solution is determined uniquely only if 

dV 
both V and -  are given for all values of x such that  x l, when   = 

(where I denotes the length of the string). 

Physical intuitions will usually indicate the nature of the boundary 
conditions which are necessary to determine a solution of a differential 
equation uniquely ; but the existence theorems which are necessary from 
the point of view of the pure mathematician are usually very tedious and 
difficult*. 

18"3. A general solution of Laplace s equation . 

It is possible to construct a general solution of Laplace's equation in the 
form of a definite integral. This solution can be employed to solve various 
problems involving boundary conditions. 

Let V  x, y, z) be a solution of Laplace's equation which can be expanded 
into a power series in three variables valid for points of  x, y, z) sufficiently 
near a given point  Xf , yo, z ). Accordingly we write 

x = Xq + X, y=zy  + Y, z = Zo+ Z; 

and we assume the expansion 

V = c(o + a X + biY + CiZ + a.. X- + b. Y- + c  Z  

+ 2cLYZ+2e,ZX + 2f,XY+..., 

it being supposed that this series is absolutely convergent whenever 

\ X'f+\ Yr + \ Z'  a, 

where a is some positive constant :]:. If this expansion exists, V is said to 
be analytic at (xo, yo, •S'o). It can be proved by the methods of §§ 3*7, 4'7 

* See e.g. Forsyth, Theory of Functions (1918), §§ 216-220, where an apparently simple 
problem is discussed. 

t Whittaker, Math. Ann. lvii. (1902), p. 333. 

* The functions of applied mathematics satisfy this condition. 



18-3] THE EQUATIONS OF MATHEMATICAL PHYSICS 389 

that the series converges uniformly throughout the domain indicated and 
may be differentiated term-by-term with regard to X, Y or Z any number of 
times at points inside the domain. 

If we substitute the expansion in Laplace's equation, which may be 
written 

d'V a F d V   
dX-' dY-' dZ-' ' 

and equate to zero (§ 373) the coefficients of the various powers of X, Y 
and Z, we get an infinite set of linear relations between the coefficients, 
of which 

a.2 4- 62 -I- c. = 
may be taken as typical. 

There are  n(n — l) of these relations* between the   n + 2) n + l) 
coefficients of the terms of degree n in the expansion of V, so that there 
are only  (n+2)(n + 1) - n (n - 1) = 2n -f 1 independent coefficients in 
the terms of degree n in V. Hence the terms of degree n in V must be 
a linear combination of 2n+l linearly independent particular solutions of 
Laplace's equation, these solutions being each of degree n in X, Y and Z. 

To find a set of such solutions, consider (Z + iX cos u 4- iFsin m)" ; it is 
a solution of Laplace's equation which may be expanded in a series of sines 
and cosines of multiples of u, thus : 

n n 

2 gm X, Y, Z) COS mu+ 2 h,, (X, Y, Z) sin. mu, 

m = 111 = 1 

the functions g iX, Y, Z) and h   X, Y, Z) being independent of u. The 
highest power of Z in gm  X, Y, Z) and A,,  (X, F, Z) is Z"'"  and the former 
function is an even function of Y, the latter an odd function; hence 
the functions are linearly independent. They therefore form a set of 
2?i -f- 1 functions of the type sought. 

Now by Fourier's rulef (§ 9'1'2) 

Trgm  X, Y, Z)= j  Z + iX cos u + i Fsin u)" cos mudu, 

irhyn  X, Y, Z)= \  Z + iX cos u -H iFsin uY sin mudu, 

J —77 

* If a ,s,j (where r + s + t = n) be the coefficient of A''T Z' in V, and if the terms of degree 
  . d-V d V dW   , . ., 

" ~ WV "*" Wr- "*" dZ  arranged primarily in powers of X and secondarily in powers of F, 

the coefficient fl .s.t <loes not occur in any term after Z'— -r Z< (or ZT*-   if r = or 1), and 
hence the relations are all linearly independent. 

t 27r must be written for tt in the coefficient of g   X, Y, Z). 



390 



THE TRANSCENDENTAL FUNCTIONS [CHAP. XVIII 



and so any linear combination of the 2/; + 1 solutions can be written in the 
form 



/: 



 Z + iX COS i( + iY sin i()' fn  u) du, 

where / (w) is a rational function of e*". 

Now it is readily verified that, if the terms of degree n in the expression 
assumed for V be written in this form, the series of terms under the integral 
sign converges uniformly if \ X\ \   + \ Y Jr\ Z  be sufficiently small, and so 
(§ 4*7) w e may write 

V =\    Z+iX cos u + iY sin ?()"/  (w) du. 

• -TT n-i) 

But any expression of this form may be written 

F = / F Z - iX cos u + i Y sin u, u) du, 

J —n 

where i  is a function such that differentiations with regard to X, Y or Z 
under the sign of integration are permissible. And, conversely, if F be any 
function of this type, V is a solution of Laplace's equation. 

This result may be written 



J —TT 



on absorbing the terms — 2 ,1 — * o cos m — I'yo sin m into the second variable; 
and, if differentiations under the sign of integi-ation are permissible, this 
gives a general solution of Laplace's equation ; that is to say, every solution 
of Laplace's equation which is analytic throughout the interior of some 
sphere is expressible by an integral of the form given. 

This result is the three-dimensional analogue of the theorem that 

V=f x- iy)+g x-ii/) 
is the general solution of 

ox  oy- 

[NoTE. A distinction has to be drawn between the primitive of an ordinary diflFerential 
equation and general integrals of a i)artial differential equation of order higher than the 
first*. 

Two apparently distinct primitives are always directly transformable into one another 

by means of suitable relations between the constants; thus in the case of ;T"f +y = 0, we 

can obtain the primitive Csin  x + t) from A cos .r + 5 sin .r by defining C and e by the 
equations Csin e = A, C'cos f = B. On the other hand, every solution of Laplace's equation 
is expressible in each of the forms 



/: 



f x cos < +// sin   + iz, t) dt, I g (y cos ?< +    sin u + ix,  ) du ; 

r ' ' — TT 



* For a discussion of general integrals of such equations, see Forsyth, Theory of Differential 
Equations, vi. (1906), Ch. xii. 



18 "31] THE EQUATIONS OF MATHEMATICAL PHYSICS 391 

but if these are known to be the same solution, there appears to be no general analytical 
relation, connecting the functions / and g, which will directly transform one form of 
the solution into the other.] 

Example 1. Shew that the potential of a particle of unit mass at (a, b, c) is 
1 [ " du 



2tt J -TT  z — c) + i  x - a) cos u + i y — b) sin u 
at all points for which z> c. 

Example 2. Shew that a general solution of Laplace's equation of zero degree in 
X, y, z is 

/ log (.rcos <+j/sin il + jj)  (<)(:/i, if I g t)dt = Q. 

Express the solutions -  — ; and log-; — \   in this form, where r- = x' +y' - z . 

Example 3. Shew that, in the case of the equation 

p  + g;  = x+y 

( where jo =  , q =  ), integrals of Charpit's subsidiary equations (see Forsyth, Differential 

Equations, Chap, ix.) are 

(i) p --x=y-q - = a, 

(ii) p = q + a?. 

Deduce that the corresponding general integrals are derived from 

(i) z =  , :c + af+l y-af+F a)\ 
Q =  x + af- y-af + F'  a) j' 

(ii) Az = \ (. +y)3 + 2a2  x-y)-a   x -y)-  + G a)\ 
= Aa x-y)-Aa  x->ry)-'  + G' [a) j' 

and thence obtain a diflferential equation determining the function O (a) in terms of the 
function E a) when the two general integrals are the same. 

18'31. Solutions of Laplace s equation involving Leg endive functions. 
If an expansion for V, of the form assumed in § 18'3, exists when 

Xq = y  Zq    ) , 
we have seen that we can express F as a series of expressions of the type 

I (z + ix cos u + iy sin u)"' cos mudu, (z + ix cos u + iy sin uy  sin inudu, 

J —TT J -TT 

where n and m are integers such that  m n. 

We shall now examine these expressions more closely. 
If we take polar coordinates, defined by the equations 

x= r sin 6 cos (f), y = r sin 6 sin (/>, z = r cos 6, 



392 THE TRANSCENDENTAL FUNCTIONS [CHAP. XVIII 

we have 

I  z + ix cos u + iy sin w)" cos mu du 

J — ]T 

= ?•" I  cos   + i sin 6 cos (  —  )j" cos mudu 

rn-<t> 
= ?'" (cos + i sin   cos yjr]" cos 7?l (</) + ylr) rf-v/r 

•Z — TT — I  

= ?'"  cos   + I sin   cos -v  " cos m  (p +  fr) dy\ r 

J —It 

= ?•" COS m(f>  cos d + i sin cos i  "  cos m'yjrd'yjr, 

J -TT 

since the integrand is a periodic function of -  and 

(cos   + z sin 6 cos i/r)" sin  ii/r 

is an odd function of yjr. Therefore (§ 15"61), with Ferrers' definition of the 
associated Legendre function, 

rir 27ri"' . n ! 

1 (z + ix cos u + iy sin  )" cos ??m c?ii = 7  -;' r Pn'  (cos  ) cos md). 

j -  c / (-. j + m) ! 

Similarly 



/: 



(  + iv cos i/ + 1?/ sin k)" sin ?/it<c?w = 7  -! r Fn '  (cos  ) sin md>. 

'  (71 + m) I 



Thei efore 7'' P,i"  (cos 6) cos 7?i  aiic? r ''Pn'  (cos  ) sin 7/i</> are polynomials 
in X, y, z and are particular solutions of Laplace's equation. Further, by 
§ 18"3, every solution of Laplace's equation, which is analytic near the origin, 
can be expr essed in the form 

F = i r'  \ AnFn (cos (9) + I (yl " ' cos m<  + Bn '''  sin m ) P "  (cos 6)1 . 

Any expression of the form 

AnPn (cos  ) + i ( n<"" cos ? </) + 5 <'   sin m(f>) P,r (cos  ), 

m = \ 

where w is a positive integer, is called a surface harmonic of degree 7i ; 
a surface harmonic of degi-ee n multiplied by ?'" is called a solid harmonic 
(or a spherical Jiarmonic) of degree n. 

The curves on a unit sphere (with centre at the origin) on which P  (cos 6) vanishes 
are n parallels of latitude which divide the surface of the sphere into zones, and so P  (cos d) 

is called (see S 15*1) a zonal harmonic ; and the curves on which . mtb . Pn"  (cos 6) vanishes 

are n—7n parallels of latitude and 2m meridians, which divide the surface of the sphere 
into quadrangles wh(> e angles are right angles, and so these functions are called tesseral 
harmonics. 



18 '4] THE EQUATIONS OF MATHEMATICAL PHYSICS 393 

A solid harmonic of degree n is .evidently a homogeneous polynomial of degree n in 
X, y, z and it satisfies Laplace's equation. 

It is evident that, if a change of rectangular coordinates* is made l;)y rotating the axes 
about the origin, a solid harmonic (or a surface harmonic) of degree n transforms into 
a solid harmonic (or a suz'face harmonic) of degree n in the new coordinates. 

Spherical harmonics were investigated with the aid of Cartesian coordinates by 
W. Thomson in 1862, see Phil. Trans. (1863), pp. 573-582, and Thomson and Tait, 
Treatise on Natural Philosophy i. (1879), jip. 171-218 ; they were also investigated 
independently in the same manner at about the same time by Clebsch, Journal fur Math. 
LXi. (1863), pp. 195-262. 

Example. If coordinates r,  ,   are defined by the equations 

1 1 

a; = rcos , ?/ = (/'— 1)- sin   cos 0, 3 = (/'2- 1)- sin  sin( , 

shew that P,/" (r) P '" (cos 6) cos wi0 satisfies Laplace's equation. 

18'4. The solution of Laplace's equation which satisfies assigned boundary 
conditions at the surface of a sphere. 

We have seen (§ 18"31) that any solution of Laplace's equation which 
is analytic near the origin can be expanded in the form 



w = ( 

+ i (  <" ' cos m(f) + 5 " ' sin m(f>) P,,"* (cos 6) \ 

m = l J 

and, from § 3*7, it is evident that if it converges for a given value of r, 
say a, for all values of 6 and (f) such that O  tt, — tt  tt, it converges 
absolutely and uniformly when r < a. 

To determine the constants, we must know the boundary conditions 
which V must satisfy. A boundary condition of frequent occurrence is 
that F is a given bounded integrable function of 6 and (f), say f(0, (f>), on 
the surface of a given sphere, which we take to have radius a, and V is 
analytic at points inside this sphere. 

We then have to determine the coefficients An, -4 <'"', 5,i""* from the 
equation 



f d,( )= S a'M Pn (cos  )+ 2 (  ' ' cos ?n</) + £ " ' sin m</))P,;'  (cos 6')   

rt = i m = l J 

Assuming that this series converges uniformly f throughout the domain 

multiplying by 

P '  (cos  )  °%i<i, 
sm   

* Laplace's operator  — , + ;r— , + ; -5 is invariant for changes of rectangular axes. 
ox  dij- oz- 

t This is usually the case in physical problems. 



394 THE TRANSCENDENTAL FUNCTIONS [CHAP. XVIII 

integrating term-by-term (§ 4'7) and using the results of §§ 15'14, 1551 on 
the integral properties of Legendre functions, we find that 

fid', 4>') P,r (cos 6') cos m<f>' sin e'dd'dcf ' = 7ra  -   . )   " ', 

Cn fir 9 (yx -1- 171, "  t 

A 6*', 6') P '  (cos d') sin wt<i)' sin O'dO'd ' = -jra"   =- . ) ( ; 5 <' ', 

f f V( '' <l>') Pn (cos  ') sin e'dB'dcf)' = lira"  5-?—.   . 
Therefore, when r < a, 
F(r, e,4>)=l ?  f-)" f  f7( ', f )]p,(cos )P (cos ') 

+ 2 i (!i:i ; p  m (cos 6) Pn'  (cos  ') COS m (6 - d)')  sin O'dO'd )'. 

The series which is here integrated term-by-term converges uniformly 
when r <a, since the expression under the integral sign is a bounded 
function of 6, 0', (f), (f>', and so (§ 4*7 ) 

47rF(r, 6*, 0) = T 1 /(0', cf>') I (2n -h 1) f-VlPn (cos  ) P  (cos  ) 

+ 2 i  y' —  Pn'"" (cos d)Pn''' (cos d') cos 7n(4>-(f>')\ sin e'de'd(f>'. 
m=i(n + m)l J 

Now suppose that we take the line (6, (f>) as a new polar axis and let 
( 1'.  1') be the new coordinates of the line whose old coordinates were (6', < ') ; 
we consequently have to replace Pn (cos 6) by 1 and P, !  (cos 6) by zero ; and 
so we get 

47rF(r, e, (f>)=r t  f(0', </)') t (2n + l)(-Y Pn(cos0,')sme,'de,'d<f>,' 

J -TT J W=0 \ \  V 

= r rf(0''<f>') i (2n+l)(-Y Pn(cose,')sme'dO'd(f>'. 

If, in this formula, we make use of the result of example 23 of Chapter xv 
(p. 332), we get 

 ' '  J-nJo \ r'-2arcos6,'+a-)  
and so 

47r F(r, e, </)) 

, r f" f(0', <i>') sin e'dd'dd)' 

= a(a- — r')\ I  - — —n- 

J -nJ [r- — 2ar  cos 6 cos  ' -f sin 6 sin  ' cos ((  — </>)  + a"]  

In this compact formula the Legendre functions have ceased to appear 
explicitly. 



18 '5] THE EQUATIONS OF MATHEMATICAL PHYSICS 395 

The last formula can be obtained by the theory of Green's functions. For properties 
of such functions the reader is refei'red to Thomson and Tait, Natural Philosophy., 
§§ 499-519. 

[Note. From the integrals for V (r, 6, cf)) involving Legendre functions of cos  j' and 
of cos , cos ' respectively, we can obtain a new proof of the addition theorem for the 
Legendre polynomial. 

For let 

Xn  \&', (t>') =Pn (COS d ) - |P  (cos 6) P  (cOS 6')  

+ 2 2  — -f; P - (COS 6) P,;" (cos 6') cos ra (0 - < ') , 
and we get, on comparing the two formulae for V  r, 6, (f)), 

0= r f'/iff, < ') 2 (2 +i) (-Xxn (d\ 4>') siD e'd\&d ,'. 

J -TT J n=0 \ / 

If we take/( ', 0') to be a surface harmonic of degree n, the term involving r" is the only 
one which occurs in the integrated series ; and in particular, if we take/( ', 4>') = Xn i 'i 0')> 
we get 

' Xni6',fp')   'id'de'd(t)' = 0. 



-n J 

Since the integrand is continuous and is not negative it must be zero ; and so 
Xn ff, (f)') = 0; that is to say we have proved the formula 

Pn (cos  i') = Pn (cos 6) Pn (COS 6') + 2 2 ) '- . Pn'" (COS 6) Pn"" (COS 6') COS 711  (j>- (f)'), 

m-1  n + M) ! 
wherein it is obvious that 

cos  ]' = cos 6 cos 6' + sin 6 sin 0' cos (0 - < '), 
from geometrical considerations. 

We have thus obtained a physical proof of a theorem proved elsewhere* (§ I5"7) by 
purely analytical reasoning.] 

Example 1. Find the solution of Laplace's equation analytic inside the sphere /•=1 
which has the value sin 36 cos <  at the surface of the sphere. 

[ s r Pgi (cos 6) cos (f) - irPji (cos 0) cos 0.] 

Example 2. Let fii r, 6, (f)) be equal to a homogeneous polynomial of degree n 
in  , y, z. Shew that 

l" I fnla, G, (f)) Pn  cos 6 con d' + siudamd' cos  (f) -4>')  a'  sin 6 d0dcf> 
J -  J o' 

- ; /.( . '< )- 

[Take the direction (6', < ') as a new polar axis.] 

18*5. Solutions of Laplace's equation luhich involve Bessel coeffi,cients. 
A particular case of the result of § 18'3 is that 



Qk(z+ixcosu+iysmu, gQg f)iuclu 

is a solution of Laplace's equation, k being any constant and m being any 

integer. 

* The absence of the factor ( - )'"  which occurs in § 1-5-7 is due to the fact that the functions 
now employed are Ferrers' associated functions. 



396 THE TRANSCENDENTAL FUNCTIONS [CHAP. XVIII 

Taking cylindrical-polar coordinates  p, (f), z) defined by the equations 
.   = p cos  , y = p sin 4>, 
the above solution becomes 

 kz I gikpcos iu-4>) cos  (   dii = gkz i gikpcos V g g .   (v + (f)) . clv 

J —IT J — JT 

= 2e*  I 6**'" °  " cos ?HV cos m<pdv 

Jo 

= 2e*  cos (mcj)) j e'*'" °  " cos mvdv, 

Jo 
and so, using § 17*1 example 3, we see that 'Itti '  e''  cos (m(f)) . Jm(f 'p) is a 
solution of Laplace's equation analytic near the origin. 

Similarly, from the expression 

T 

where m is an integer, ive deduce that 27rz'" e*  sin (??i</)) .    (kp) is a solution 
of Laplace's equation. 

18 "SI. The periods of vibration of a uniform membrane*. 

The equation satisfied by the displacement V at time t of a point  x, y) of a uniform 
plane membrane vibrating harmonically is 

'dx  cy- c- Ct- ' 

where c is a constant depending on the tension and density of the membrane. The 
equation can be reduced to Laplace's equation by the change of variable given by z = cti. 
It follows, from § 18'5, that expressions of the form 

'  sm   sin 
satisfy the equation of motion of the membrane. 

Take as a particular case a drum, that is to say a membrane with a fixed circular 
boundary of radius R. 

Then one possible type of vibration is given by the equation 

r=t    kp) cos m<\ i cos ckt, 
provided that F=0 when p = R ; so that we have to choose k to satisfy the equation 

J,  kR)=0. 
This equation to determine /  has an infinite number of real roots (§ 17'3 example 3), 
 1,  2j  3j ••• S3,y. A possible type of vibration is then given by 

r=/ , (f-'rp) cosmcf) cos ckrt  r= 1, 2, 3, ,..). 

This is a periodic motion with period 1iTJ ckj.) ; and so the calculation of the periods 
depends essentially on calculating the zeros of Bessel coefficients (see   17'9). 

• Euler, Novi Covim. Acad. Petrop. x. (1764) [published 1766], pp. 243-260; Poisson, Mem. 
de I'Academie, viii. (1829), pp. 357-570; Bourget, Ann. de I'Ecole nvrm.sup. iii. (1866), pp. 55-95. 
For a detailed discussion of vibrations of membranes, see also Rayleigh, Theory of Sound, 
Chapter ix. 



18 "5 1-1 8 '61] THE EQUATIONS OF MATHEMATICAL PHYSICS 397 

Example. The equation of motion of air in a circular cylinder vibrating per- 
pendicularly to the axis OZ of the cylinder is 

dx  dy'  c' dt- ' 
V denoting the velocity potential. If the cylinder have radius R, the boundary condition 

dV 
is that n— =0 when p = R. Shew that the determination of the free periods depends on 

finding the zeros of J,,/ (C) == 0- 

18'6. A general solution of the equation of wave motions. 
It may be shewn* by the methods of § 183 that a general solution of 
the equation of wave motions 

dx  dy'' dz- c- di? 



IS 



F = I I f(x sin u cos v + y sin u sin v + z cos n + ct, u, v) dudv, 



where /is a function (of three variables) of the type considered in § 18"3. 

Regarding an integral as a limit of a sum, we see that a physical 
interpretation of this equation is that the velocity potential V is produced 
by a number of plane waves, the disturbance represented by the element 

f(x sin u cos v + y sin u sin v + z cos u + ct, u, v) 8u Bv 
being propagated in the direction (sin u cos v, sin u sin v, cos u) with velocity c. 
The solution therefore represents an aggregate of plane waves travelling in 
all directions with velocity c. 

18'61. Solutions of the equation of wave motions which involve Bessel 
functions. 

We shall now obtain a class of particular solutions of the equation of 
wave motions, useful for the solution of certain special problems. 

In physical investigations, it is desirable to have the time occurring by 
means of a factor sin ckt or cos ckt, where k is constant. This suggests that 
we should consider solutions of the type 

/•tt rn 

y  =, \ I giA;(a;sinttcost;+2/sinMsiuv+2cosM+cti f (u i)\ dudv 

J -TT J 

Physically this means that we consider motions in which all the elementary waves 
have the same period. 

Now let the polar coordinates of (x, y, z) be (r, 6, (f>) and let (o), yjr) be the 
polar coordinates of the direction (u, v) referred to new axes such that 
the polar axis is the direction (ff, (f)), and the plane -v/r = passes through 
OZ; so that 

cos CO = cos 6 cos u + sin d sin u cos ((f) — v), 

sin II sin ((f) — v) = sin w sin yjr. 

* See the paper previously cited, Math. Ann. lvii. (1902), pp. 342-345, or Messenger of Mathe- 
matics, XXXVI. (1907), pp. 98-106. 



398 THE TRANSCENDENTAL FUNCTIONS [CHAP. XVIII 

Also, take the arbitrary function /(w, y) to be ;S,j (w, y) sin i  where Sn 
denotes a surface harmonic in u, v of degree ?i ; so that we may write 

Sn(u, V) = Sn  0, 4>] CO, yjr), 

where (§ 18*31) Sn is a surface harmonic in co, -sjr of degree n.   
We thus get 

V = e ' ' f r I ' e''""°"" Sn  6, < \ w, y\ r) sin a> dw df. 

Now we may write (§ 18 "SI) 

Sn (d, (i>; 0 ,y r) = An e,(f>). Pn (COS Oj) 

+ S    "'" (0, <f>) cos myfr + Bn'"'   0, (f)) sin myfr] P "' (cos co), 

m = l 

where An  6, </>),  n'"'*  6, <p) and 5 ""'  6, 0) are independent of yjr and co. 
Performing the integration with respect to -yjr, we get 

F= 27re'' '   An  0, 4>) f"e '*"° "P  (cos to) sin codco 

Jo 

= 27re' '-  A n  0,(ii)j e''"-'  P,, (/i) djM 

= 27r. - '* An (  , ( ) J'  e'* '  -  (f.  - l)n d , 

by Rodrigues' formula (§ loll); on integrating by parts ti times and using 
Hankel's integral (§ 17 '3 corollary), we obtain the equation 

 = o;r '"" n  0, 4>)  ikrY I e'>"-  (1 - /x )" c yu 

Zi • 'It \ J 1 

= (27r)  iV'* '   kr) - -J  +   (kr) An (6, c >), 
and so F is a constant multiple of e' ' h-'   Jj \  \ i(kr) An(0, (f>). 

Now the equation of wave motions is unaffected if we multiply x, y, z 
and t by the same constant factor, i.e. if we multiply r and t by the same 
constant factor leaving Q and   unaltered ; so that An B, <f)) may be taken 
to be independent of the arbitrary constant k which multiplies r and t. 

Hence lim e**-''' r ~ H' ~ " ~   /  , i (kr) An  0, 9) is a solution of the equation 

A:-*0   

of wave motions; and therefore r' An(0, <f>) is a solution (independent of t) 
of the equation of wave motions, and is consequently a solution of Laplace's 
equation ; it is, accordingly, permissible to take An (0, <f>) to be any surface 
harmonic of degree n ; and so we obtain the result that 



r -   Jn+r ih-) Pn''' (cos 0)   md>   ckt 
 2   sm   sm 

is a particular solution of the equation of wave motions. 



I8"61lj THE EQUATIONS OF MATHEMATICAL PHYSICS 399 

18*611. Application of § 18-61 to a physical problem. 

The solution just obtained for the equation of wave motions may be used in the 
following manner to determine the periods of free vibration of air contained in a rigid 
sphere. 

The velocity potential T' satisfies the equation of wave motions and the boundary 

dV 
condition is that - =- =0 when ?  =  , where a is the radius of the sphere. Hence 

or   

~\ -, ,i.r, , /iv COS , COS , 

Y-r *J  ,.(/-/•)  ,™ cos<9) . wd) . ckt 
"+-   " sm   sin 

gives a possible motion if Z* is so chosen that 

This equation determines k; on using § 17'24, we see that it may be written in 
the form 

tan ka =    ka)  

where f  (ka) is a rational function of ka. 

In particular the radial vibrations, in which V is independent of 6 and < , are given by 
taking n = 0; then the equation to determine k becomes simply 

tan ka = ka ; 
and the pitches of the fundamental radial vibrations correspond to the roots of this 
equation. 

REFERENCES. 

J. Fourier, La theorie analytique de la Chaleur. (Translated by A. Freeman.) 

W. Thomson and P. G. Tait, Natural Philosophy. (1879.) 

Lord Rayleigh, Theory of Sound. (London, 1894-1896.) 

F. PoCKELS, Uber die partielle Diferentialgleichung /\ u + khi. = 0. (Leipzig, 1891.) 

H. BuRKHARDT, Eiitivickelungen nach oscillirenden Funktionen. (Leipzig, 1908.) 

H. Bateman, Electrical and Optical Wave-motion. (1915.) 

E. T. Whittaker, History of the Theories of Aether and Electricity. (Dublin, 1910.) 

A. E. H. Love, Proc. London Math. Soc. xxx. (1899), pp. 308-321. 

H. Bateman, Proc. London Math. Soc. (2), i. (1904), pp. 451-458. 

L. N. G. FiLON, Philosophical Magazine (6), vi. (1903), pp. 193-213. 

H. Bateman, Proc. London Math. Soc. (2), vii. (1909), pp. 70-89. 

Miscellaneous Examples. 

1. If V be a solution of Laplace's equation which is symmetrical with respect to OZ, 
and if y=f z) on OZ, shew that if f  \ be a function which is analytic in a domain of 
values (which contains the origin) of the complex variable f, then 



'=i r -   + i(.r2+j/2)5cos0 rf(  
  / 



at any point of a certain three-dimensional region. 

Deduce that the potential of a uniform circular ring of radius c and of mass M lying in 
the plane XO T with its centre at the origin is 



  r [  +    +   ( ' + '/)  cos (/) 2] - I d<t>. 

7!" y 



400 THE TRANSCENDENTAL FUNCTIONS [CHAP. XVIII 

2. If r be a solution of Laplace's equation, which is of the form e""*jP(p, 2), where 

(p, (f>, z) are cylindrical coordinates, and if this solution is approximately equal to 

p"'e""*/(i) near the axis of z, where /(f) is of the character described in example 1, 
shew that 



 = r ( i+ 1 ) r (*) /  ' "   ''''  ' '"""  ' - (DougaU.) 



3. If ?  be determined as a function of .r, y and z by means of the equation 

A.v+By + Cz=\, 
where A, B, C are functions of ?< such that 

shew that (subject to certain general conditions) any function of   is a solution of 

Laplace's equation. 

(Forsyth, Messenger, xxvii. (1898), pp. 99-118.) 

4. A, B are two points outside a sphere whose centre is C. A layer of attracting 
matter on the surface of the sphere is such that its surface density (Tp at P is given by 
the formula 

apCc AP.BP)-\ 

Shew that the total quantity of matter is unaffected by varying A and B so long as 

CA . CB and ACB are unaltered ; and prove that this result is equivalent to the theorem 

that the surface integral of two harmonics of different degrees taken over the sphere 

IS zero. 

(Sylvester, Phil. Mag. (5), il. (1876), pp. 291-307.) 

5. Let V (.r, y, z) be the potential function defined analytically as due to particles 
of masses X + iy., X - z/x at the points (a + ia', b + ib', c + ic') and (a — ia', b - ib', c — ic') 
respectively. Shew that V  x, y, z) is infinite at all points of a certain real circle, and 
if the point (.r, y, z) describes a circuit intertwined once with this circle the initial 
and final values of V x,y, z) are numerically equal, but opposite in sign. 

(Appell, Math. Ann. xxx. (1887), pp. 155-156.) 

6. Find the solution of Laplace's equation analytic in the region for which a<r<iA, 
it being given that on the spheres /•=a and r = A the solution reduces to 

2 c P (cos(9), 2 C;P (cos<9), 
respectively. 

7. Let 0' have coordinates (0, 0, c), and let 

Pdz=6, P0'Z=6', PO = r, PO' = r'. 
Shew that 

P  cos6') \  P  cosd ) . .s cPn i(cos ) jn + 1) Qi + 2) c'' P  +    cos d) 
( Y" f   I (n I 1) - 1 (  " ) 1  n + l) n + 2) r 'P, oo,d) 1 

"  ~  / Un + l V" / pn + 2 '  2! C""  T...|, 

according as r>c or r<c. 

Obtain a similar expansion for /"P ' (cos  ). (Trinity, 1893.) 

8. At a point (r, 6, (p) outside a uniform oblate spheroid whose semi-axes are a, b and 
whose density is p, shew that the potential is 



)a-b 



;\ \  \  m  Pj (cos 0) m  P  (cos 6) 



-...]. 



3/- 3.5 r  b.l r° 

where ni  — a--b- and ?•>? .. Obtain the i otential at points for which r<m. 

(St John's, 1899.) 



THE EQUATIONS OF MATHEMATICAL PHYSICS 401 

9. Shew that 

eirco,e=(  )i I in [271 + 1) r-lP   cos 6) J +i(r). 

n =   

(Bauer, Journal fiir Math, lvi.) 

10*. Shew that if x ±iy = h cosh (| + irj), the equation of two-dimensional wave motions 
in the coordinates   and  ? is 

a|7 + 5  =   (   h- 1 - cos2 rj) -  . (Lame.) 

IL Let X —  c + r cos 6) cos (p, y =  c + r cos 6) sin cp, 2 = /-sin ; 

shew that the surfaces for which /•, d, (f) respectively are constant form an orthogonal 
.system; and shew that Laplace's equation in the coordinates / , \&, is 

— - r(c + '/-cos )  +-;  (c + ?-cos(9)-; [+—  7., =0. 

vr \   a- j r 06 \ j- ' cd \ c + rcosdd(f>- 

(W. D. Xiven, J/essenger, x.) 

12. Let P have Cartesian coordinates  x, ?  z) and polar coordinates (;•, 6, (f)). Let 
the plane POZ meet the circle x'  + t/' = k-, s = in the points a, y; and let 

aPy = (o, log (Pa/Py) = a. 
Shew that Laplace's equation in the coordinates o-, a>, (p is 

8 f sinho- bV] d  j sinho- \  871 1  -0- 

da- [cosh o- — cos a da j ca (cosh a — cos o) 8w j sinh o- (cosh cr — cos w) 8( 2 ' 
and shew that a solution is 

V= (cosh o" — cos 0))  cos /iQ) cos ??i0 P (cosh o-). 

(Hicks, Phil. Trans. CLXXii. p. 617 et seq.) 

13. Shew that 

00  -hn i r-j- r  

 R  + p -2Rpcoscl) + c-)-h= 2 I dk I e-"" J,   I- p)e' ' ''°"' cos rauclu, 

m=0   y J -T 

and deduce an expression for the potential of a particle in terms of Bessel functions. 

14. Shew that if a, b, c are constants and X, /x, v are confocal coordinates, defined as 
the roots of the equation in e 

a2 + e' 6-  + e'  6-2 + 6 ' 
then Laplace's equation may be written 

Ax(M-) xf  + M(-X)| A/J +A.(X-M)  A. ]=0, 

where A;, = V (aHX) (i'  + X) (c-  + X) . 

(Lame.) 

* Examples 10, 11, 12 and 14 are most easily proved by using Lame's result (Journal de 
VEcole Polyt. xiv. cahier 23 (1834), pp. 191-288) that if (X, /ul, v) be orthogonal coordinates for 
which the line-element is given by the formula  dx)'' + (8yf +  5zy =(H d\ y  + (H2diM)'- +  H 5v)' , 
Laplace's equation in these coordinates is 

d fH.H cV\ 8 / H. H, dV\ 8 / H,H., dV\ \   



'Hi- 



d\ V Hi a,\ J dfjL\ H., d/jiJ di'\ Hs dv 
A simple method (due to W. Thomson, Camh. Math. Journal, iv. (1845), pp. 83-42) of proving 
this result, by means of arguments of a physical character, is reproduced by Lamb, Hydro- 
dynamics (1916), § 111. Analytical proofs, based on Lame's proof, are given by Bertrand, 
Traite de Calcul Differentielle (1864), pp. 181-187, and Goursat, Coiirs d'Analyse, i. (1910), 
pp. 155-159, the last proof being appreciably the simplest. Another proof is given by Heine, 
Theorie der Kugelfunctionen, i. (1878), pp. 303-306. 

W. M. A. 26 



402 THE TRANSCENDENTAL FUNCTIONS [CHAP. XVIII 

15. Shew that a general solution of the equation of wave motions is 

r= F  .v coH 6 +   iiin \$ + iz, >/ + iz siu 6 + ct cos B, B)dd. 

J —IT 

(Bateman, Proc. London Math. Soc. (2) l. (1904), p. 457.) 

16. If r=/(.r, y, z, t) be a solution of 

d  ct cx' cy'  dz- ' 
prove that another solution of the equation is 

17. Shew that a general solution of the equation of wave motions, when the motion is 
independent of < , is 

/ / (2 + ip cos e, ct + p sin 6) dO 

f' /""   , / a -   z + ct cos d\  ,  , ,. , 

+ arc suih —  r—   ) F (a, 6) dBda, 

J (1 .' -TT \ p sin   / ' ' 

where p, (f>, z are cylindrical coordinates and a, h are arbitrary constants. 

(Bateman, Proc. London Math. Soc. (2) i. (1904), p. 458.) 

"  18. If r=/(.r, y, z) is a solution of Laplace's equation, shew that 

Y  1  / r -a  r +gg az \ 

~( \ , )i-' V2(.'--i '  i x-iyy x-iy) 

is another solution.   

(Bateman, Proc. London Math. Soc. (2) vii. (1909), p. 77.) 

19. If U=f x, y, z, t) is a solution of the equation of wave motions, shew that 
another solution is 

rj  1 .( X y \ r2-l rHl \ 

z-ct-'Xz-ct'' z-ct"   z-cty 2c z-ct))' 

(Bateman, Proc. London Math. Soc. (2) vii. (1909), p. 77.) 

20. If l=x-iy, m = z + hv, n = . : - + y'  + z - + iv% 

\ = j:+iy, fji = z-uv, p= — \, 
so that l\ + 7nfi + nv=0, 

shew that any homogeneous solution, of degree zero, of 

cHT cH[ dH/ c U Q 
dx'  dy  cz  dw  

satisfies  +Z£+ =0- 

dldX dmdfi dndv ' 

and obtain a solution of this equation in the form 

j' , b, c \ 

U',  ', y' ) 

where   = ( -c) (f-a),. ? /Li = (c-a) (f-6), np = (a- b) ((-c). 

(Bateman, Proc. London Math. Soc. (2) vii. (1909), pp. 78-82.) 



THE EQUATIONS OF MATHEMATICAL PHYSICS 403 

21*. If (r, 6, (f)) are spheroidal coordinates, defined by the equations 

jr = c (r  + iy sin   cos< , y=c (r2-|-l) sin  sin( , z=crcosd, 

where x, y, z are rectangular coordinates and c is a constant, shew that, when n and m are 
integers, 

('"   /.  cos  + y sin -f i2\ cos ,   (n — in)\ \   ,.     ,   cos 

Pn\  —]   mtdt = 27ry—  —  P,r tr)Pn'"  cos 6) . md). 

j -n- \ c /Sin  n+m)l     "   sin   

(Blades, Proc. Edinburgh Math. Soc. xxxiii.) 
22. With the notation of example 21, shew that, if 2 =t= 0, 

/    /cPcos  + y sin  +i'2\ cos , ,  (n — m)l  , . s -r, /   cos 
Qn  - — )   intdt = 277 )--— -fj § '   ir) P '  cos e) . m4>. 
-r \ c J Sin (n+m) i  "   -' "   ' sin   

(Jeffery, Proc. Edinburgh Math. Soc. xxxiil.) 

* The functions introduced in examples 21 and 22 are known as internal and external 
spheroidal harmonics respectively. 



26—2 

