\chapter{The Hypergeometric Function} 

141. The hypergeometric series.

We have already (§ 2*38) considered the hypergeometric series*

1 + - . , ( + l) ( + l) a(a + l)(a+2)6(6+l)(6 + 2) 3 l.c 1.2.c(c + l)
"* 1.2.3.c(c + l)(c + 2) " ••

from the point of view of its convergence. It follows from § 2-38 and
§ 5-3 that the series defines a function which is analytic when i 2 j
< 1.

It will appear later (§ 14-53) that this function has a branch point
at 2 = 1 and that if a cutf (i.e. an impassable barrier) is made from
+ 1 to + along the real axis, the function is analytic and one-valued
throughout the cut plane. The function will be denoted by F a, b;c;z).

Many important functions employed in Analysis can be expressed by
means of hypergeometric functions. Thus|

(1 +2)" = F(-n, : ;-z),

log l + z) = zF(l, l;2;-z),

e'=\ im F l,0;l;zl0).



Example. Shew that d



F a, b; c; z) = ~F a + \, 6 + 1; c + 1; 2).



1411. The valuel of F(a, b; c; I) when E(c-a-b)>0.

The reader will easily verify, by considering the coefficients of
.2;'* in the

* The name was given bj- Wallis in 1655 to the series whose Kth term
is a + b a + 2b ... a + n-l)b . Euler used the term hypergeometric in
this sense, the modern use of the term being apparently due to Kummer,
Journal fiir Math. xv. (1S36).

t The plane of the variable z is said to be cut along a curve when it
is convenient to consider only such variations in z which do not
involve a passage across the curve in question ; so that the cut may
be regarded as an impassable barrier.

X It will be a good exercise for the reader to construct a rigorous
proof of the third of these results.

§ This analysis is due to Gauss. A method more easy to i-emember but
more difficult to justify is given in § 14-6 example 2.



282 THE TRANSCENDENTAL FUNCTIONS [CHAP. XIV

various series, that if \$ ir < 1, then c c-l- 2c-a-b-l)x]F (a, b : c
; x) + (c - a) c - b) xF (a, b ; c + I ; x)

= c(c-l) l-x)F(a,b;c-l;x)

= C(C -l)jl+ (Un-Un-i)x'>'l,

where Un is the coefficient of " in i''(a, 6 ; c — 1 ; x).

Now make x—>l. By § 3*71, the right-hand side tends to zero if

1+2 (un — tin-i) converges to zero, i.e. if ?i,j— >0, which is the
case when =i

R c- a - 6) > 0.

Also, by § 2*38 and § 3'71, the left-hand side tends to

c(a+b-c)F a,b;c;l) + c-a)(c- b) F(a, b;c+l;l) under the same condition
; and therefore

Repeating this process, we see that

F(a, b: c;l) = i Yl ) —— -, ( F(a, b:c + 7n; 1)

  ' '\ n=o c + n) c-a-b + n)\ ' ' '

= ] lira n ) . , f [ lim F(a, b;c + m; 1),

if these two limits exist.

But ( 12'13) the former limit is tt r-Fr, rz, if c is not a negative

• T c — a) 1 (c— 6) °

integer; and, if Un(ci, b, c) be the coefficient oi x in F (a, b; c;
x), and

m > I c I , we have

00

\ F a, b] c + m; l) — l\ \ 2 \ un(a, b, c + m)\

n = l

00

  X Un (ja,, ib\, m — \ c )

n=l

' Q,b I "

< , 2 Mn (I a i -h 1, I 6 I + 1, 7?l -f- 1 - I C I).

m-\ c\ n=o

Now the last series converges, when ??i > ic| 4- a -f j l — 1, and is
a positive decreasing function of m; therefore, since [m — '\ c\ \ }~
—>0, we have

lim F(a, b: c + m ; 1)= 1 ;

and therefore, finally,

   , , r(c)r(c-a-6)



14*2, 14-3] THE HYPERGEOMETRIC FUNCTION 283

14"2. The differential equation satisfied by F (a, b; c: z).

The reader will verify without difficulty, by the methods of § lO'S,
that the hypergeometric series is an integral valid near = of the
hypergeometric

equation*

   ~ d? + (c - (a + + 1) - - \&' = ;

from § 10"3, it is apparent that every point is an 'ordinary point' of
this equation, with the exception of 0, 1, x , and that these are '
regular points.'

Example. Shew that an integral of the equation

i

z< F a- a, 6 + a; a- + 1; z).

14'3. Solutions of Riemann's P-equation by hypergeometric functions.

In § 10"72 it was observed that Riemann's differential equation f

dru j] -a -a' 1- -/3' l-y-y\ du dz- \ z— a z—b z — c ) dz

\ aa' (a-b)(a-c) jB' h - c) b - a ) \ 77 (c-a)(c-6) z— a z — b z — c

u



 z — a) z — b) z — c)



= 0.



by a suitable change of variables, could be reduced to a
hypergeometric equation ; and, carrying out the change, we see that a
solution of Riemann's equation is

fz-ay /z-c\ y 3 r./ 1 ' ( -a)(c-\&)]

provided that a — a' is not a negative integer ; for simplicity, we
shall, throughout thi section, suppose that no one of the exponent
differences a — a', /3 — , 7 — 7' is zero or an integer, as (§ 10"32)
in this exceptional case the general solution of the differential
equation may involve logarithmic terms ; the formulae in the
exceptional case will be found in a memoir :J: by Lindelof, to which
the reader is referred.

Now if a be interchanged with a', or 7 with 7', in this expression, it
must still satisfy Riemann's equation, since the latter is unaffected
by this change.

* This equation was given by Gauss.

t The constants are subject to the condition a + a' + + ' + y + y' =
l.

J Acta Soc. Sclent. Fennicae, xix. (1893). See also Klein's
lithographed Lectures, Veher die hypergeometrische Funktion (Leipzig,
1894).



284 THE TRANSCENDENTAL FUNCTIONS [CHAP. XIV

We thus obtain altogether four expressions, namely,

'z-aX /z-c\ y f . , , - . , , ,\ c-h) z-a)\



M, =



(z-ay /z-c\ y p , ry, -, > c-b) z-a)]

 ==(.— i) U— ft) J | + /3 + 7', o + +t; l+°-'''; (o-a)(.-6) j-

which are all solutions of the differential equation.

Moreover, the differential equation is unaltered if the triads (a, a',
a), (yS, y8', h), (y, y', c) are interchanged in any manner. If
therefore we make such changes in the above solutions, they will still
be solutions of the differential equation.

There are five such changes possible, for we may write

 b, c, a\, c, a, b], a, c, b], [c, b, a], [b, a, c]

in turn in place of [a, b, c], with corresponding changes of a, a, /S,
(3', 7, 7'.

We thus obtain 4 x = 20 new expressions, which with the original four
make altogether twenty-four particular solutions of Riemann's
equation, in terms of hypergeometric series.

The twenty new solutions may be written down as follows :



u. =



z-by fz-aY ( , , , a-c)(z-b)\ iP'j/8 + 7 + a, /3 + 7'+a; l + yS-/3'; )
, (

z-b\ \ ' [z-ay t q, , o' , ', . 1 , /D' o. (a-c)( -\&)]



Ua =



 s=('— ) ('- Y Fy+y + oc, '+y'+a-l + /3'-/3- ,. .

\ z - cj \ z - cj [ " a-b) z- c)

(z-b fz — a-y \ ( , \ , , , (a — c)(z — b)

\ z — cl \ z — c \ (a —b) z — c)

i /3+7+a, /3+7+a ; l + /3'-/3; ~ '-\

\ Z-C' \ Z-C' y I ' r- I ) r- r- ' ( \ J - c)\

(z-c\ y (z-b\ \ , o -. , b-a) z- c)]

" "=U- ) U- J |7 + ,7+a-+ ;l + y-7; \ ; \ |

U,,= l - )'( - r F\ y a + ',y c '; I + 7-7'; j H \ z-aj \ z-al [' ' "
' ' ' (l,-c)(z-a)\

   = (lllfy7l Yf f/'+o + ZS'. 7'+a+/3'; 1 + 7-7; P - \ z-aj \ z-a) ' "
' ' ' > (i,-c) z- a)



14-4] THE HYPERGEOMETRIC FUNCTION 285



Z ~ Cj



a,.=



MlQ —



FL + y + ',a + y' + '; I + a - a ; l' '\ \ ' ' ''\ \ , [ b-a) z-c)

/ -cy'/2-a\ j ' , /o' , 1, ' a-b)(z-c))

.T b) U— 6J |t+/3 + ,7+ +< : 1+7-7; \ ;( \ |,

\ z - a J \ z-a) [ ' (c-b)(z- a) '

— / i c — o)(z — a))

1 V - / \ z-aj ( ' " I r ' (c-6)( -a)J '

By writing 0, 1 - C, , B, 0, 6'- -5, a; for a, a', yg, S', y, y,

/ — AU — respectively, we obtain 24 solutions of the hypergeoraetric

equation satisfied hy F A, B; G ; x).

The existence of these 24 solutions was first shewn by Kuinmer*.

14"4. Relations betiueen particular solutions of the hyper geometric
equation.

It has just been shewn that 24 expressions involving hypergeometric
series are solutions of the hypergeometric equation ; and, from the
general theory of linear differential equations of the second order,
it follows that, if any three have a common domain of existence, there
must be a linear relation with constant coefficients connecting those
three solutions.

If Ave simplify n- , u.,, ih, ih] u , u g-, u i, W22 in the manner
indicated at

* Journal fUr Math. xv. (1836), pp. 39-83, 127-172. They are obtained
in a different manner in Forsyth's Treatise on Differential Equations,
Chap. vi.



' •22 = ,

\ z — aj \ z



286 THE TRANSCENDENTAL FUNCTIONS [CHAP. XIV

the end of § 14-3, we obtain the following solutions of the
hypergeometric equation with elements A, B, G, x:

y, =F A,B;C-x),

y = (- xy- 'FiA -C+\,B-G+1;2-C;x),

y,=(l-xr-''- F C-B,C-A;G;xl

y = (\ ccY-'' l - xf- - 'Fil -B,l-A;2-G;x),

y,r = F A,B; A + B - G + 1 ; 1 - x),

y = xf- -''F G-B, G-A; G-A-B-hl; l-x),

y = (\ xY 'FiA, A-G+\;A-B + \; x-%

y = (- x)- F B,B-G + l]B-A + l; x-'). If 1 arg(l — a;) I < TT, it is
easy to see from § 2-53 that, when j .r | < 1, the relations
connecting y , y.2, y y must be y y-i, y2= rj , by considering the
form of the expansions near c ; = of the series involved.

In this manner we can group the functions Wj, ... z/24 into six sets
of four*, viz. u-i, v-i, Mi3, Wis; Un, Hi, u , iiis; ii , u , M21, u
', Wg, u , U22, M24 j '"9? Wji, t<i7> ' 19 j Wjo, U12, u-18, '' ho,
such that members of the same set are constant multiples of one
another throughout a suitably chosen domain.

In particular, we observe that Ui, u , u s, u are constant multiples
of a function which (by §§ 5*4, 2"o3) can be expanded in the form



 z-aY\ l+ i en(2-a)4



when \ z — a\ is sufficiently small ; when arg z — a) is so restricted
that (z — a)* is one-valued, this solution of Riemann's equation is
usually written P'"*. And P'"'; P' ', P' '; P* ', P* * are defined in
a similar manner when



z — a



\ z — b\, \ z — c\ respectively are sufficiently small.



To obtain the relations which connect three members of separate sets
of solutions is much more difficult. The relations have been obtained
by elaborate transformations of the double circuit integrals which
will be obtained later in § 14-61 ; but a more simple and singularly
elegant method has recently been discovered by Barnes ; of his
investigation we shall give a brief account.

14"5. Barnes contour integrals for the hypergeometric function .

Consider r—. ~ -- p/ , . — — '(-zyds,

273-1 j\ xj r(c + s)

where |arg(— -z)] <7r, and the path of integration is curved (if
necessary) to ensure that the poles of r(a + 5)r(6 + 6'), viz. s = — a
— n, —h—n n = 0, 1, 2, . . .),

* The special formula

FiA,l;C;.) F( C-A,l;C; ),

which is derivable from the relation counectiug Mj with i3, was
discovered in 1730 by Stirling, Methodus Differential is, prop. vii.

t Piuc. London Math. Soc. (2), vi. (1908), pp. 141-177. References to
previous work on similar topics by Pincherle, Mellin and Barnes are
there given.



14-5]



THE HYPERGEOMETRIC FUNCTION



287



lie on the left of the path and the poles of r(— 5), viz. s = 0, 1, 2,
..., lie on the right of the path *.

From § 13'6 it follows that the integrand is

Oils j +''-c-i exp - arg - z) . I (s) - 7r\ I s)\ \ ]

as s— ►CO on the contour, and hence it is easily seen (| 5"32) that
the integrand is an analytic function of z throughout the domain
defined by the inequality I arg 2\ TT — S, where B is any positive
number.

Now, taking note of the relation F (— 5) F (1 + s) = — tt cosec s-rr,
consider

T(a+s)r(b+s) iri-zy



If IttiJ c



ds,



27rij(7 F(c + s) F(l +s) sinsTT

where C is the semicircle of radius iV + - on the right of the
imaginary axis

with centre at the origin, and N is an integer.

Now, by § 13"6, we have

r(a+ )F(6 + ) -n-i-zy . a+,\ ,\ ,) (- y T(c + s)r l+s) sinsTT ' sin
stt

as N—> 00 , the constant implied in the symbol being independent of
arg s when s is on the semicircle ; and, if 6 = f j\ \ + j e''* and j
| < 1, we have

(— zy cosec sir = exp j ( iV + .j j cos 6 log | | — ( lY + j sin arg
(— z)

-(iV + )7r|sm ||]

exp I i\ \ + . ) cos 6' log I I - (iV + ] 8 1 sin ( 1 11

exp|2- iV +. )logj

exp|-2- 8(i\ r + i)|



=



  I 6* k 7 TT,







1 I zii 1



Hence if log | | is negative (i.e. \ z\ < 1), the integrand tends to
zero sufficiently rapidly (for all values of 6 under consideration) to
ensure that



/.



'0 as iV— oo



Now



J -ooi [J ~xi J C J (iV+i) i)



by Cauchy's theorem, is equal to minus 2'Tri times the sum of the
residues of the integrand at the points s = 0, 1,2, ... N. Make N—>cc
, and the last

* It is assumed that a and b are such that the contour can be drawn,
i.e. that a and b are not negative integers (in which case the
hypergeometric series is merely a polynomial).



288 THE TRANSCENDENTAL FUNCTIONS [CHAB. XIV

three integrals tend to zero when | arg (— ) | tt — S, and \ z \ < 1,
and so, in these circumstances,

the general term in this summation being the residue of the integrand
at

S = 71.

Thus, an analytic function namely the integral under consideration)
exists throughout the domain defined by the inequality j arg | < tt,
and, when \ z\ < 1, this analytic function may he represented by the
sei'ies



V



r(a + n)r(b + n) r c + n) .n !



The symbol F(a, h; c; z) will, in future, be used to denote this
function divided by r( .)r(6)/r(c).

14'51. The continuation of the hyper geometric series.

To obtain a representation of the function F (a, b; c; z) in the form
of series convergent when j 2 | > 1, we shall employ the integral
obtained in § 14'5. If -D be the semicircle of radius p on the left of
the imaginary axis with centre at the origin, it may be shewn* by the
methods of § 14'o that

V a + s)V h + s)T -s)



Jd r c + s)



as p— x , provided that j arg — z) \ < tt, \ z \ > 1 and p—><xi in
such a way that the lower bound of the distance of D from poles of the
integi and is a positive number (not zero).

Hence it can be proved (as in the corresponding work of § 14"5) that,
when I arg —z)\ < 7r and \ z', > 1,

1 r r a + s)r(b + s)T -s)

\ V r(a+7i)r(l -c + a + n) sin ( c- a - n) tt /\ x- -h ~ rZo r (1 + w)
r (1 - 6 + ft + n) cos n-rr sin (b - ft - n) it

  T(b + n)r l-c + b + n) sin (c - 6 - ?i) tt f \

n=o r (1 + n) r (1 — ft + 6 + n) cos nir sin (ft — 6 — w) tt

the expressions in these summations being the residues of the
integrand at the points s = - a- n, s = - b — n respectively.

It then follows at once on simplifying these series that the analytic

* In considering the asymptotic expansion of the integrand when | s [
is large on the contour or on D, it is simplest to transform T (a +
s), V (b + s), T c + s) by the relation of § 12-14.



14-51, 14*52] THE HYPERGEOMETRIC FUNCTION 289

continuation of the series, by which the hypergeometric function was
originally defined, is given by the equation

r(c) ""' '- ' r(a-c) -(- r' - ( >i- + ;i- + )

1 0 — c) where arg (— z)\ < tt.

It is readily seen that each of the three terms in this equation is a
solution of the hypergeometric equation (see § 14"4).

This result has to be modified when a — 6 is an integer or zero, as
some of the poles of r a + s)r h + s) are double poles, and the
right-hand side then may involve logarithmic terms, in accordance with
1 43.

Corollary. Putting h = c, we see that, if | arg - z)\ < n,

r (a) z)-" = -. f "' r a+s) r ( -s) ( - zyds, 'ZniJ \ xt

where (1 - )~<'- -l as 2-*.0, and so the value of [ arg(l - z) \ which
is less than tt always has to be taken in this equation, in virtue of
the cut (sec i:; 14'1) from to -f-oo caused by the inequality | arg( -
z) \ < n.

14*52. Barnes' lemma that, if the 'path of integration is curved so
that the poles of T y — s)T b — s) lie on the right uf the path and
the poles of T (a + s) r (/3 + s) lie on the left*, then

Write / for the expression on the left.

If C be defined to be the semicircle of radius p on the right of the
imaginary axis with centre at the origin, and if p-s-oo in such a way
that the lower bound of the distance of C from the poles of r (y - s)
r (5 - s) is positive (not zero), it is readily seen that

T a + s)T \& + s)T y-s)T b-s) = - ''' '

= 0[..' - +>+ - exp -2 1/( )l J, as ] s |- -oo on the imaginary axis
or on C.

Hence the original integral converges ; and / - -Oasp-a-oc , when (a +
/3-|-y-|-S- 1)<0.

Thus, as in § 14'5, the integral involved in 7 is - 2ni times the sum
of the residues of the integrand at the poles oiT y — s)T b — s) ;
evaluating these residues we gett

/= i r( a + y+%)rO + y + ? ) TT I V a + b + n)T + b + n)



a=oV n + ) T + y-8 + 7l) sin(S-y)7r ,,=o T (?i+l) r (1 -f-S-y + w)
sin(y-S)7r*

* It is supposed that a, /3, y, 5 are such that no pole of the first
set coincides with any pole of the second set.

t These two series converge (§ 2-38).

W. M. A. 19



290 THE TRANSCENDENTAL FUNCTIONS [cHAP. XIV

And so, using the result of 12"14 freely, by § 14-11 :

\ 7rr(l-a-/3-y -5) r T (a + d) T (3 + 5) r(a + y ) rO + y) )

sin(y-8)7r |r(l-a-y)r(l-/3-y) T (1 - a -8)~r (1 - -8)J

r(a+y)r( + y)r(a + g)r( + a) f r , ro ,

- r(a+/J + y + 8)sm(a + + y + 5).sin(y-8). " ( + ) ' " ('

-sin (a + S) 77 sin 04-S)7r|.

But 2sin(fi+y)7rsiu(/3 + y)7r-2sin(a + S)7rsin (3 + S)7r

= cos (a — /3) TT - cos (a + /3 + 2y) tt - cos (a — jS) tt + cos (a +
+ 28) tt

= 2sin(y-8)7r sin(a+/3 + y + S)7r.

Therefore Vja + y)n ym-\ + ) V )

 (a + + y + S)

which is the required result ; it has, however, only been prt)ved when

 (o + /3 + y + 8-l)<0;

hut, )y the theory of analytic continuation, it is true throughout the
domain through which both sides of the equation are analytic functions
of, say, a: and hence it is true for all values of a, /ii, y, 8 for
which none of the poles of r (a + s) r (/3 + s), qua function of s,
coincide with any of the poles of r (y - s) r (8 - s).

Corollary. Writing s + k., a — k, \$-k, y + k, 8 + k in place of s, a,
/3, y, 8, we see that the result is still true when the limits of
integration are —; ,• + cx) i, where k is any real constant.

14 "53. The connexion between hypergeometrie functions of z and of \
—z. We hav§ seen that, if | arg ( - s) | < tt,

r(c) ' ' ' 2ni j \ x,- T c + s) "• '

= . \ \ r-.\ T a t)Y h-irt)T s-t)V c-a-h-t)dt\

'' TTi J \ x i l TTl J ~k—x i )

.T c-a)T c-b) ' by Barnes' leinma.

If k be so chosen that the lower bound of the distance between the s
contour and the t contour is positive (not zero), it may be shewn that
the order of the integrations* may be interchanged.

Carrying out the interchange, we see that if arg (1 -z) be given its
principal value,

T(C'-a)T c-h)T a)r b)F a, b; c; 2)/r (c)

= s— / r a + t)r b+t)r c-a-b~t) —. r s-t)ri-s)(~zYds dt

 T lJ-k-Jii 'ZtvI J -jci

1 f-k+oci

= r—. I r a + t)r b+t)r c~a-b-t)r -t) i-zy dt.

2ni J -A-K/

* Methods similar to those of § 4-51 may be used, or it may be proved
without much difficulty that conditions established by Bromwich,
Infinite Series, % 177, are satisfied.



14'53, 14 6] THE HYPERGEOMETRIC FUNCTION 291

Now, when I arg (1 -k) \ < Stt and | I -z\ < 1, this last integral may
be evaluated by the methods of Barnes' lemma (§ 14-52) ; and so we
deduce that

r c-a)r c-b)T a)rib)F a,b;c;z)

= r c)r a)r b)T c-a-b) F a, b ; a + b-c + l ; l-z)

+ r c)r c - a) r (c-b) r a + b - c) I -zy-'"-" F c- a, c-b; c-a-b + l;
z),

a result which shews the nature of the singularity of F a, b ; c ; z)
at z=.

This result has to be modified if c — a — 6 is an integer or zero, as
then

Y a-- t) V b->r t) T c - a -b - t)T -t)

has double poles, and logarithmic terms may appear. With this
exception, the result is valid when | arg —z)\ < tt, | arg (1 - 2) ; <
tt.

Taking | s | < 1, we may make 2 tend to a real value, and we see that
the result still holds for real values of 2 such that < 2 < 1.

14"6. Solution of Riemann's equation hy a contour integral.

We next proceed to establish a result relating to the expression of
the liypergeometric function by means of contour integrals.

Let the dependent variable u in Riemann's equation (§ 10"7) be
replaced by a new dependent variable defined by the relation

u = (z - aY z - bf (z - c)y I.

The differential equation satisfied by / is easily found to be

d f 1 + a a' l+ -§' 1-f - 7) dl dz- \ z — a z — h z — c ] dz

( g + /3 + 7) (g + + ry + 1) + Sa (g + /?' + 7' - 1)1 z — a) z — 0) z
— c)

which can be written in the form

Q(z) -[ X-2)Q' z)+R z)]f

+ li ( - 2) (X - 1 ) (/' (z) - (X-DR (z)] 1 = 0,

where / \=l-ot- -j = a' + ' + y',

iQ(z) z-a)(z-h) z-c),

(R(z) = X a' +/3 + ry) z-b) z-c).

It must be observed that the function / is not analytic at x , and
consequently the above differential equation in / is not a case of the
generalised hypergeometric equation.

We shall noiu sheiu that this differential equation can be satisfied
by an integral of the form

1=1 t- ay'+ +y- (t - 6)-+ '+y-i (t - c)-+ -y -1 z - )- - -y dt, J c

jirovlded that C, tJie contour of integration, is suitably cJiosen.

19—2



292 THE TRANSCENDENTAL FUNCTIONS [CHAP. XIV

For, if we substitute this value of / in the differential equation,
the con- dition* that the equation should be satisfied becomes

j (t- ay+ +y-' t - by+ '+y-' t - cY+p+y'-' z - t)-''- -y--Kdt = 0,

J c where

ir = (\ - 2) |q (z) + t- z) Q' (z) + l(t- zj Q" z)



+ (t-z) R(z) + (t-z)R'(z)]

= ( 2) Q (t) -it- zy] -it-z) [R (t) -a- zy s ( ' + /3 + 7)1

= - (1 + a + /3 + 7) ( - a) ( - 6) (i - c)

+ S (a + /9 + 7) ( - ) ( - c) t - z).

f dV It follows that the condition to be satisfied reduces to -7- dt =
0, where

J c dt

v=(t- ay+ +y (t - by+ '+y (t - cy+ +y (t - )-(i+-+p+v .

The integral / is therefore a solution of the differential equation,
when C is such that V resumes its initial value after t has described
C.

Now

V= t- ay'+ +y-' (t - by+ '+y-' (t - cy+ +y-' [z - t)-''- -y U,

where U = (t- a) t -b) t- c) (z - ty\

Now Z7 is a one-valued function of t ; hence, if C be a closed
contour, it must be such that the integrand in the integral / resumes
its original value after t has described the contour.

Hence finally any integral of the type

(z-ay(z-by(z - c)y \ (t-ay+y+''-\ t-b)y+''+ '-' t-cy+ y'-' z-ty''~ -y
dt,

JC

•where C is either a closed contour in the t-plane such that the
integrand resumes its initial value after t has described it, or else
is a simple curve such that V lias the same value at its termini, is a
solution of the differential equation of the general hypergeometric
function.

The reader is referred to the memoirs of Pochhammer, Math. An)i. xxxv.
(1890), pp. 495-526, and Hob.son, Phil. Trans. 187 a (1896), pp.
443-531, for an account of the methods by which integrals of this type
are transformed so as to give rise to the relations of § 14-51 and
14-53.

Example 1. To deduce a real definite integral which, in certain
circumstances, represents the hypergeometric series.

• The iliEferentiations imder the sign of integration are legitimate
(§ 4-2) if the path C does not depend on z a3d does not pass through
the points a, b, c, z ; if C be an infinite contour or if C passes
through the points a, b, c or z, further conditions are necessary.



14-61] THE HYPERGEOMETRIC FUNCTION 293

The hypergeometric series F a, b; c; z) is, as already shewn, a
solution of the differential equation defined by the scheme

r 00 1 \

P a zy

\ \ —c b c- a- b ] If in the integral

which is a constant multiple of that just obtained, we make 6 -x
(without paying attention to the validity of this process), we are led
to consider

/ t'''' t-\ y- - t-zydt.

Now the limiting form of V in question is

and this tends to zero at t= 1 and = cc , provided R (c) > R(b)>0.

We accordingly consider / t"~'= t — Vf-''-' t-z)~" dt, where z is not*
positive and greater tlum 1.

In this integral, write t = u~ ; the integral becomes

/ ? ''-' (1-jt)' -''-' (l-M2)- 0??<. .'

We are therefore led to expect that this integral may be a solution of
the differential equation for the hypergeometric series.

The reader will easily see that if R (c) > R b)> 0, and if arg u = arg
(1 - m) =0, while the branch of l-uz is specified by the fact that
(1-m2)- 1 as m- 0, the integral just found is

T b)r c-b)

  ( , \&; c;z .

This can be proved by expandingf (1 — s)-a in ascending powers of z
when | 3 | <1 and using § 12 "41.

Example 2. Deduce the result of § 14-11 from the preceding example.

14'61. Determination of an integral which represents i "\

We shall now shew how an integral which represents the particular
solution i ' (§ 14-3) of the hypergeometric difterential equation can
be found.

We have seen (§ 14-6) that the integral

l= z- df [z~bf z - cy ( t-a)P+y+- '-\ t-b)y+<'+P'- (t-c)'' +
+y'-Ht-z)-<'~ -ydt

satisfies the difl:erential equation of the hypergeometric function,
provided C is a closed contour such that the integrand resumes its
initial value after t has described C. Now the singularities of this
integrand in the plane are the points a, b, c, z; and after describing
the double circuit contour (§ 12'43) symbolised by (6 + , c-h, 6 — , c
— ) the integi-and returns to its original value.

* This ensures that the point t — ljz is not on the path of
integration, t The justification of this process by § 4'7 is left to
the reader.



294 THE TRANSCENDENTAL FUNCTIONS [CHAP. XIV

Now, if z lie iu a circle whose centre is , the circle not containing
either of the points h and c, we can choose the path of integration so
that t is outside this circle, and so \ z — a\ < i\ t — a\ for all
points t on the path.

Now choose arg(2 — a) to be numericall ' less than it and arg(i — 6),
ai*g(2-c) so that they reduce to* arg(a — 6), arg(a-c) when z- a\ fix
arg( — a), arg (< — 6), arg (< — c) at the point iV at which the path
of integration starts and ends ; also choose arg t - z) to reduce to
arg t — a) when z- a.

Then (z-bf = (a-bf |l +/3 (j ' ) + ...j- ,

(.-c)V=( -o)' l+ (i ?) + ...l, and since we can expand (t-z)~' ~°~'
into an absolutely and uuifornily convergent series

we may expand the integral into a series which converges absolutely.

Multiplying up the absolutely convergent series, we get a series of
integer powers of z — a multiplied by (s — a) . Consequently we must
have

We can define P '''\ P \ P '\ F y\ P ' by double circuit integrals in
a similar manner.

14'7. Relations hetiueen contiguous hypergeometric functions.

Let P z) be a solution of Riemann's equation with argument z,
singularities a, h, c, and exponents a, a', /3, j3', y, 7'. Further
let P(z) be a constant multiple of one of the six functions P * ,
P'"'', P' ', P' \ P<y , P't*. Let Pi+i,m-i(z) denote the function
which is obtained by replacing two of the exponents, I and m, in P 2)
hy I + 1 and m — 1 respectively. Such functions P;+i, i\ i (z) are
said to be contiguous to P (z). There are 6 x 5 = -30 contiguous
functions, since I and m may be any two of the six exponents.

It was first shewn by Riemannf that the function P(z) and any tivo of
its contiguous functions are connected hy a linear relation, the
coeffi,cients in which are polynomials in z.

There will clearly be 5 x 30 x 29 = 435 of these relations. To shew
how to obtain them, we shall take P z in the form

F z) = z- ay (z - hY (z - c)y f t- ay+y+' '- (t - 6)v+' + -

J c

(t - cY+ +y'-' (z - ty- -y dt,

where C is a double circuit contour of the type considered in § 14*61.

* The values of arg (a-b), arg (a - c) being fixed.

t Abh. der k. Ges. der Wiss. zu Gottingen, 1857; Gauss had previously
obtained 15 rehitions between contiguous hypergeometric functions.



147] THE HYPERGEOMETRIC FUNCTION 295

First, since the integral round G of the differential of any function
which resumes its initial value after i has described C is zero, we
have

= [ I- (t - ay+ +y (t - by ' y-Ht - cY+ +y'-' t - zy- - -y] dt. J c dt

On performing the differentiation by differentiating each factor in
turn,

we get

(a' + + y)P + (a + ' + y-l) Pa-+,, '-, + (a + (3 + y'-l) Pa'+i,y-i

\ (a + /3 + 7) p

Considerations of symmetry shew that the right-hand side of this
equation can be replaced by

These, together with the analogous formulae obtained by cyclical
inter- change* of (a, a, a) with (6, /3, ') and (c, 7, 7'), are six
linear relations connecting the hypergeometric function P with the
twelve contiguous functions

Pa+], '-l, -I p+i y'—i, -/y+i,a\ l, -ta-],v'-l5 -Lfi+i a'-l, y~\, '~\
i -I a.'+l, '— 1) -I a'4-],Y'\ i, i '4.),-y'\ i, -t j3'+l, a'-l > ''
y'+I, a'- 1 ) -*7'+l, '— 1-

Next, writing t — a = (t- h) + (6 — ), and usingf Pa'\ i to denote the
result of writing a' — 1 for a' in P, we have

P = P..\ :,,.+:- (6- )P.-:.

S-imilarly P = P \ i,y'+i + (c - a) Pa'-i-

Eliminating Pa-i from these equations, we have

 c-h)P + (a - c) Pa-\ :, vi + (b- a) Pa'-.,y-,, = 0. This and the
analogous formulae are three more linear relations con- necting P with
the last six of the twelve contiguous functions written above.

Next, writing ( — ) = ( — a) — ( — a), we readily find the relation

P = P +:, v'-i - ( - )" ' ( - f y ( - 0

Jc which gives the equations

( \ a)-i P - ( - 6)- P +,, Y--: = ( - h)- [P- z- c)- Py+, a'-i

= Kz - c)-' [P- z- a)-' Pa+i, \ i .

* The interchange is to be made only in the integrands ; the contour C
is to remain unaltered.

t Pa'-l is not a function of Riemann's type since the sum of its
exponents at a, h, c is not unity.



296 THE TRANSCENDENTAL FUNCTIONS [CHAP. XIV

These are two more linear equations between P and the above twelve
contiguous functions. *

We have therefore now altogether found eleven linear relations between
P and these twelve functions, the coefficients in these relations
being rational functions of 2. Hence each of these functions can be
expressed linearly in terms of P and some selected one of them ; that
is, between P and any two of the above functions there exists a linear
relation. The coefficients in this relation will be rational functions
of z, and therefore will become polynomials in z when the relation is
multiplied throughout by the least common multiple of their
denominators.

The theorem is therefore proved, so far as the above twelve contiguous
functions are concerned. It can, without difficulty, be extended so as
to be established for the rest of the thirty contiguous functions.

Corollary. If functions be derived from P by replacing the exponents
a, a, /3, /3', y, y

by a+p, a' + g, /3 + r, jS'+s, y + t, y' + x, where p, q, r, s, f, u
are integers satisfying the

relation

p- -q + r + s + t + u=0,

then between P and any two such functions there exists a linear
relation, the coefficients in which are polynomials in z.

This result can be obtained by connecting P with the two functions by
a chain of intermediate contiguous functions, writing down the linear
relations which connect them with P and the two functions, and from
these relations eliminating the intermediate contiguous functions.

Many theorems which will be established subsequently, e.g. the
recurrence-formulae for the Legendre functions (§ 15-21), are really
cases of the theorem of this article.

REFERENCES. C. F. Gauss, Ges. Werke, in. pp. 12.3-163, 207-229. E. E.
KuMMER, Journal fur Math. xv. (1836), pp. 39-83, 127-172. G. F. B.
RiEMANN, Ges. Math. Werke, pp. 67-84. E. Papperitz, Math. Ann. xxv.
(1885), pp. 212-221. S. PiNCHERLE, Rend. Accad. Lincei (4), iv.
(1888), pp. 694-700, 792-799. E. Y. Barxes, Proc. London Math. Soc.
(2), vi. (1908), pp. 141-177. Hj. Mellin, Acta Soc. Fennicae, xx.
(1895), No. 12.

Miscellaneous Examples.

1. Shew that

F a, h + \; c; z)-F a, b; c; z) = — F a + l, h + l; c + l; z).

2. Shew that if o is a negative integer while /3 and y are not
integers, then the ratio F(a, IB; a + /3 + l -y; \ - .v)-7-F a, ; y ;
A') is independent of x, and find its value.



THE HYPERGEOMETRIC FUNCTION 297

dP d P

3. If P iz) be a hypergeometrie function, express its derivates -,-
and -- linearly in

dP

terms of P and contiguous functions, and hence find the linear
relation between P, -j- ,

d'-P and -TV , i.e. verify that P satisfies the hypergeometrie
difterential equation.

4. Shew that i j, j ; 1 ; 4 (1 -z)) satisfies the hypergeometrie
equation satisfied by jP(|, I ; 1 ; z). Shew that, in the left-hand
half of the lemniscate \ z z) | = j, these two functions are equal ;
and in the right-hand half of the lemniscate, the former function is
equal to F \, | ; 1 ; 1 —2).

5. \ i Fu, =F a + ,h; c; .r), / \ =i (a - 1, h; c; x), determine the
15 linear relations with polynomial coefficients which connect F a, h;
c; x) with pairs of the six functions Fa , Fa-, F,,, F,\ , F, , F,\ .
(Gauss.)

6. Shew that the hypergeometrie equation

x(x-l/J - y-(a+ + l)x ' + a i/ =

is satisfied by the two integrals (suppo.sed convergent)

[\ \ -\ l-z)y- -' l-xz)-''dz J

and ['/-i(l-2)"-i' l-(l- )2 ~"rf2.

.' "

7. Shew that, for values of x between and 1, the solution of the
equation

is AF a,i ; i; \ --Ixy-l + B l-2x) F h a + l), h fi + l); |; l-2xy- ,

where A, B are arbitrary constants and F a, •,y;x) represents the
hypergeometrie series.

(Math. Trip. 1896.)

8. Shew that

Jim|F( ,/3,y,..)- J (-) ,;r(y-a)r(y- )r( )r( ) ' J

\ r(y-a- )r(y ) (y-a) (y- )

where h is the integer such that k R (a + — y)<k- .

(This specifies the manner in which the hj'pergeometric function
becomes infinite when .r- -l -0 provided that o-f |3-y is not an
integer.) (Hardy.)



9. Shew that, when i? (y - a - /S) < 0, then

-/3-y

- -l



T y)n'' -y



 "•( 4- -y)r(a)r( )

as '/i -x ; where S denotes the sum of the first n terms of the series
for F a, /3 ; y ; 1).

(M. J. M. Hill, Proc. London Math. Soe. (2), v.)



298 THE TRANSCENDENTAL FUNCTIONS [CHAP. XIV

10. Shew that, ii i/i, y-i be indeijeiident sokitions of



• 2- 1- -.



then the oreneral sohition of



is z = Ay - Byiy.i- Cyi', where J, B, C are constants.

(Appell, Comptes Rendus, xci.)

11. Deduce from example 10 that, if + ft + | = c,

 F(a h- r- r)\ \ = \ rW\ r(2c-l) - T 2a + 7>) F ia + b+n)T 2b+7i) ,,

'' ' '' r 2a)T 2b)T a + b)n=o nl T c + n)T 2c- 1+n) *' '

(Clausen, Journal fiir Math, iii.)

12. Shew that, if | | <i and j x \ —s) j <|,

F 2a, 2/3; a+/3 + *; x = F a, /3 ; a+ + l ; 4.r(l-.r) . (Kummer.)

13. Deduce from example 12 that

14. Shew that, if co = e ' ' and i (a) < 1,

i (a,3a-l; 2a; " - ) = 3 '' '* exp [K Sa - 1) | | ,

i ( ,3a-l; 2a; -co) =3 - exp W (1 -3a) M | .

(Watson, Quarterly Journal, xhi.)

15. Shew that

V 2 ) 2 -rj> -r\ j ..y V9; r(|)r('/i + f)

(Heymann, Zeitschrift fur Math, und Phys. XLiv.)

16. If il-xY+ -y F 2a,2 ;2y; x) = + Bx + C:'f- + Dx +..., shew that

i (a, /3; y+ ; ) i (y-a, y-/3 ; y+l; x)

y+l (7+l)(7 + i-) (y+*)(y + f)(7+|)

(Cayley, PAz7. Mag. (4), xvi. (1858), pp. 356-357. See also Orr, Camb.
Phil. Trans, xvii. (1899), pp. 1-15.)

17. If the function F a, ji, /3', y ; .*;, y) be defined by the
equation

i (a, ,/3', y; x,y) = \ - \ \ i' -\ \ ~xc)y-''-\ \ - ux)-\ l-uy)- '
du,

r (a) r (y - a) j

then shew that between F and any three of its eight contiguous
functions

i (a±l), F ii± ) , F ii'±l), F y±\ \

there exists a homogeneous linear equation, whose coefficients are
polynomials in x and y.

(Le Vavasseur.)



THE HYPERGEOMETRIC FUNCTION 299

18. If y - a - iS < 0, shew that, a.s x - 1-0,

and that, if -0-/3 = 0, the corresponding approximate formula is

(Math. Trip. 1893.)

19. Shew that, when \ x\ < 1,



/•(2 + ,0 + ,a:-,0-) 1 1 fl



= \ ,r/' sin OTT sin (y - a) TT . ' — " - "V (a, /3 ; y ; ),

where c denotes a point on the straight line joining the points 0, x,
the initial arguments of v — x and of v are the same as that of .>;,
and arg (1 — j/) -*-0 as v- 0.

(Pochhammer.)

20. If, when \ arg (1 -x)\ < 2n,

K x)=—r T -s)Ti ; + s)\ \ ' xYds,

LTTl J \ y, ,

and, when | arg.*; | < 27r,

l-lZl J \ K,-

by changing the variable s in the integral or otherwise, obtain the
following relations : K (.r) = A" (1 - ), if I arg (1 - :f) I <

K l-x) = K'(x), if|argj;|<7r.

K x) = l - .r)-4 A' ( - , if i arg (1 - .r) \ < n.

K l-x) x-h K f"-— , if I arg.r | <7r.



K' x) = x 2 / ' (Ijx), if I arg.r |'< tt.

A" (1 - a;) = (1 - X) - i A" ( \ J\ - \ ") , if j arg (1 - .x-) I < -
(Barnes.)

21. With the notation of the preceding example, obtain the following
results :

11=0 [ 'I- )

2 K' x)= - l f lt— j'- - log -4log 2 + 4 (1 - 1 + -- .

when \ x\ < l, | arg .i? | < tt ; and

iT ( ) = + i ( - ) - 5 a: (l/.r) + ( - ) - 2 ' (1/:??),

when j arg — x)\ < ir, the ambiguous sign being the same as the .sign
of / (x).

(Barnes.)



300 THE TRANSCENDENTAL FUNCTIONS [CHAP. XIV

22. Hypergeometric series in two variables are defined by the
equations F, a; , ';y; .r, y) = 2 "Xn f" '"J/'S

F, (a ; /3, ' ; y, y ; ,r, y) = 2 '-, x y .



-ts a,a, , ; y; x, y)= 2 , --,-\ x 'y"



 n( n Pm Pn ,( ??l . n i y i ,1



FUa, ;y,y';.,y) .XpV ' '"'" '' where a, = a(a + l). ..(a + m — 1), and
2 means 2 2 .

?n, n m=0 =0

Obtain the differential equations

c"F. d'F ?Ft ?F

-(l-)8 +y(l-)5j;5 + r-( +e+l)- 5;j'-* -. /-,=0,

d F d F dF;

.il-o:)- +y Hy- a + + l)x]j -a F, = 0,

and four similar equations, derived from these by interchanging x with
y and a, /3, y with a', 3', y when a', ', y occur in the corresponding
series.

(Appell, Comptes Rendus, xc.) 23. If a is negative, and if

a= —V + a,

where v is an integer and a is positive, shew that



r (x) r (a)






u D (-)"( -l)(a-2) ... (a-%) , where = — - - — ' — — " ' G( -n),



" i (•*/ = — . (Hermite, Joiirnal mr Math, xcii.)

x+n \ 1 . /



24. When a < 1, shew that



T x)T a-x) I \ l Rn



T a) n=\ X + n n=\ X — a — n''

where (-) a(a + l) ... (a + .-l)

7i !

25. When a > 1, and v and a are respectively the integral and
fractional parts of a, shew that

T x)T a-x) G x)p.,\ - (?( ')p, r(a) =i .r + ?i =i . r-a — n

-G x) + + ...+ f'-'X '\ \ X — a X-a-\ .r-a-i + lj'



THE HYPERGEOMETRIC FUNCTION 301



where (?( )=(l-? (l -- ) ... (l \ . \ )

V a/ V a+l/ \ a + v — lj

ajjj \ (-)"a(a+l)... (a + -l)



It, ;

(Hermite, Journal fiir Math, xcil.) 26. If

/• (T , \ r - '(jZ + + n-l) ., x x+l)(7/ + v + n-l)( + v + n)

./ (.r, y, .)- 1 - C, - - - ,) + C.3 3,( + 1) ( + ,)( . , + i) - ... ,

where n is a positive integer and C'i, C2, ... are binomial
coefficients, shew that

f (T y , r(y)rCy- -H/ )r(.r-n;)r(y+?0

./H .y. ) Y y-x)T y + n)V :6)V x-irV n)'

(Saalschiitz, Zeitschrift fur Math. xxxv. ; a number of similar
results are given by Dougall, Proc. Edinburgh Math. Soc. xxv.)

shew that, when liib + e - ,a — ) > 0, then

i (a,a-5+l, -c + l'; S, e ; l)=2- ..A P W " " — ' ' ' ' ' r(8-
a)r(f-ia)r( + ia)r(S + e-a-l)

(A. C. Dixon, P/'oc. London Math. Soc. xxxv.)

28. Shew that, if R (a) < .if, then

; f a(a+l)... (a-l-n- l)l ,. ,,, . r(l-3a)

(Morley, Proc. London Math. Soc. xxxiv.) 2!). If

[' / '....-I (1 \ .,.)>-iy-i (1 \ )A-. (1 -a,yy--J->'dxdi/ = B(i,J,
X-, m), J i> J i>

shew, by integrating with respect to jc, and also with respect to i
that B i,j, /•, m) is a symmetric function of i+J,j + i\ /• + + in,
/n+i.

Deduce that

F a,ld,y; fi, f ; l) r (8) F ( ) T (6 + 6 - a -/3-y)

is a symmetric function of 8, f , S + e — a - /i, 8 + e — - y, 8 + e -
y - a.

(A. ('. Dixon, Proc. London Math. Soc. (2), II. (1905), jjp. 8-16. For
a proof of a special case by Barnes' methods, see Barnes, Quarterly
Journal, XLI. (1910), pp. 136-140.)

30. If

x ~ il-x f'"- d" I

/ = i (- , a + ;.; y, .r) = - - y— - W ( x) \

shew that, when n is a large positive integer, and 0< .r < 1,

  = , - (sin, )i-V(cos0)V- -icos (2u + a)( -i,r(2y-l) + o( 1 ),

where .r = sin2 .

(This result is contained in the great memoir by Darboux, "Sur
I'approxi- mation des fonctions de trfes grands nombres," Journal de
Math. (3), iv. (1878), pp. 5-56, 377-416. For a systematic development
of hyper- geometric functions in which one (or more) of the constants
is large, see Camh. Phil. Trans, xxii. (1918), pp. 277-308.)

